\section{xODE: Ordinary Differential Equations}

[Simon Hood prepared the code.  We do not yet have extensive
documentation.  See modules for discussion.  This material come from
the ode interface.

This code is not yet folded into the general m3na library architecture.
See the ``hood'' subdirectory.]

This is a Library of numerical integrators to
ordinary differential equations of arbitrary order, linear or nonlinear.
Both initial value problems and
  boundary value problems are catered for:

\subsection*{IVPs}
At present only one routine is supplied;  this is based on a 
      Runge-Kutta Order Four scheme.  This is usually good enough.  I will
      be adding a Runge-Kutta Order Six scheme, simply because I have an
      implementation I did years ago knocking around, though I don't expect
      it to be significantly better very often.  More significant might be
      the addition of a predictor-corrector scheme or something specifically
      for ``stiff'' differential equations.

\subsection*{BVPs}
Two routines are/will be supplied:  
      a ``relaxation/finite-difference''
      routine and a second based on a ``shooting'' algorithm.  The latter
      is not yet implemented --- though it solves, in principle at least, 
      the same problems as the former.  In practice, one routine may converge
      when the other does not;  furthermore, one can provide an 
      independent check of the other (assuming both converge!);  finally,
      rest assured that the shooting algorithm {\it will\/} be implemented
      in the not too distant future as I need it for my research!


