\section{NADefinition: Utilities}

\subsection*{verbosity, debug}
{\tt verbosity} controls the {\tt debug} function.
0 means don't print anything and 3 means print everything.
It is up to the user of {\tt debug} to
set {\tt level} for each call of {\tt debug}.  E.g.:

\begin {tt} \begin{verbatim}
  (*verbosity has been set to 2*)

  debug(1,ftn,"this will print");
  debug(2,ftn,"this will print");
  debug(3,ftn,"this will not print");
\end{verbatim} \end{tt}

[HGG: I use {\tt debug} because I don't have m3gdb running.]

\subsection*{Error, Err, and err}
The theory is that we carefully define the contract of the interface and
expect the caller to live up to it.  However, in practice, where
it is convenient and not a tremendous overhead we do some defensive
editing.  We also report exceptions that the caller could not
realistically have been expected to know.

We raise just {\tt Error} in this library.
It gets an {\tt Err} value, which can be
used to do case decomposition in the exception handler.

The selection of {\tt Err} values has been haphazard.  We probably should
start with the POSIX matrix error enumeration, and grow as needed.
Just haven't gotten around to looking at it.

{\tt err} is a shorthand for raising {\tt Error}.
Originally, it also printed
a message to stderr. Since that is no longer the approach, err is not
much used.


There is still an open design decision:
How to declare a set of exceptions?
\begin{enumerate}
\item One exception for each error -
This implies long RAISES lists that are hard to update.
\item One global exception for the library with a number that specifies the detailed error (current approach) -
New exceptions can only be added by extending a central module.
It's not possible to add detailed error specific information.
\item One global exception with an object parameter. -
With subclasses one can achieve a more detailed error analysis.
What information should be provided by the exception object?
Is handling of exceptions still easy?
Is this method efficient?
\end{enumerate}
