%%
%% $Id$
%%
\documentclass[12pt]{report}
\usepackage{appendix}
\usepackage{tocloft}
\input gencapltd

\setlength{\cftsubsecnumwidth}{3.5em}

\title{\bfseries\sffamily\fontsize{24.88}{30}\selectfont A Modula-Scheme System: \\ MScheme \\ Description and Reference}
\author{Mika Nystr\"om \\ {\tt mika@alum.mit.edu}}
\date{November, 2008}

\thispagestyle{empty}

\begin{document}
\def\ns{}
\ifx\DIST\undefined
\setbox0=\vbox{\verb+$Id$+}
\lfoot{\copy0}
\fi

\thispagestyle{empty}

\maketitle

\thispagestyle{empty}

\newpage
\tableofcontents
\newpage

\input system
\input usermanual
\input sstubgen
\input refmanual
\input examples

\newpage
{
\frenchspacing
\begin{thebibliography}{99}

\bibitem{sicp2} Hal Abelson and Gerald Jay Sussman with Julie Sussman.  {\it Structure and Interpretation of Computer Programs,} second edition.  Cambridge, Mass.:\ MIT Press, 1996.

\bibitem{blue_book} Adele~Goldberg and Donald~Robson.  {\it Smalltalk-80: The Language and its Implementation.}  Reading, Mass.:\ Addison-Wesley, 1983.

\bibitem{spwm3}  Greg~Nelson, ed.  {\it Systems Programming with Modula-3.}  Englewood Cliffs, N.J.:\ Prentice-Hall, 1991.

\bibitem{netobj} Andrew~Birrell, Greg~Nelson, Susan~Owicki, and Edward~Wobber.  {\it Network Objects,} digital SRC Research Report~115.  Palo Alto, Calif.:\ Digital Equipment Corporation, 1994 (revised 1995).

\bibitem{dp} Erich~Gamma, Richard~Helm, Ralph~Johnson, and John~Vlissides.  {\it Design Patterns: Elements of Reusable Object-Oriented Software.}  (Addison-Wesley Professional Computing Series.)  Reading, Mass.:\ Addison-Wesley, 1995.

\bibitem{m3tk}  Mick Jordan.  An extensible programming environment for Modula-3.  {\it Proceedings of the Fourth ACM SIGSOFT Symposium on Software Development Environments,} Irvine, Calif., 1990.  New York, N.Y.:\ Association for Computing Machinery, 1990.

\bibitem{r4rs}

\bibitem{r5rs} Richard Kelsey, William Clinger, and Jonathan Rees, eds.
Revised$^5$ Report on the Algorithmic Language Scheme.  1998.

\bibitem{jscheme} Peter Norvig.

\bibitem{Tcl} John~K.~Ousterhout.  Tcl and the Tk Toolkit.  Addison-Wesley, 1994.

\bibitem{aim-349} Gerald Jay Sussman and Guy Lewis Steele, Jr.  Scheme: an intepreter for extended lambda calculus.  MIT Artificial Intelligence Memo 349.  Cambridge, Mass.:\ Massachusetts Institute of Technology, December 1975.
\end{thebibliography}
}


\end{document}
