%%
%% MScheme Compiler Design: From Binding Optimization to Native Code
%%
\documentclass[12pt]{report}
\usepackage[margin=1in]{geometry}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{textcomp}
\usepackage{amssymb}

\lstdefinelanguage{Scheme}{
  morekeywords={define,lambda,if,cond,else,let,let*,letrec,begin,
    set!,quote,quasiquote,unquote,unquote-splicing,and,or,not,
    case,do,delay,force,map,for-each,apply,call-with-current-continuation,
    call/cc,require-modules,load,macro,unwind-protect,eval,
    current-environment},
  sensitive=true,
  morecomment=[l]{;},
  morestring=[b]",
}

\lstdefinelanguage{Modula3}{
  morekeywords={MODULE,INTERFACE,IMPORT,FROM,PROCEDURE,BEGIN,END,
    VAR,CONST,TYPE,RECORD,OBJECT,METHODS,OVERRIDES,RETURN,
    IF,THEN,ELSE,ELSIF,WHILE,DO,FOR,TO,BY,LOOP,EXIT,REPEAT,UNTIL,
    CASE,OF,WITH,TRY,EXCEPT,FINALLY,RAISE,RAISES,
    NEW,ARRAY,REF,REFANY,SET,BOOLEAN,INTEGER,REAL,LONGREAL,CHAR,TEXT,
    TRUE,FALSE,NIL,BRANDED,REVEAL,EVAL,NOT,AND,OR,IN,
    ISTYPE,NARROW,TYPECODE,ADDRESS},
  sensitive=true,
  morecomment=[n]{(*}{*)},
  morestring=[b]",
}

\lstdefinelanguage{C}{
  morekeywords={typedef,char,int,void,if,else,return,while,for,
    extern,static,struct,union,const,unsigned,signed,long,double},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[s]{/*}{*/},
  morestring=[b]",
}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!5},
  keywordstyle=\bfseries\color{blue!70!black},
  commentstyle=\itshape\color{green!50!black},
  stringstyle=\color{red!60!black},
  showstringspaces=false,
  columns=flexible,
  tabsize=2,
  xleftmargin=2em,
  framexleftmargin=1.5em,
}

\title{\bfseries\sffamily\fontsize{24.88}{30}\selectfont
  MScheme Compiler Design\\[0.5em]
  \large From Binding Optimization to Native Code Generation}
\author{Mika Nystr\"om \\ {\tt mika@alum.mit.edu}}
\date{March 2026}

\begin{document}
\maketitle
\thispagestyle{empty}

\newpage
\tableofcontents
\newpage

%======================================================================
\chapter{Introduction}
%======================================================================

The MScheme interpreter contains an optimization called the
\emph{binding optimization}, described in detail in the companion
report \emph{MScheme Internals}.  When a closure is created, the
\texttt{SchemeClosure.Init} procedure walks the body and replaces
free variable symbols with \texttt{SchemeEnvironmentBinding.T}
objects that point directly to the environment cell where each
variable lives.  At evaluation time, these binding objects are
recognized by the evaluator and resolved in constant time, bypassing
the scope-chain traversal.

This document explores the question: can this optimization be
extended into a framework for compiling Scheme closures to native
Modula-3 (or C) code?

We begin by analyzing what work the evaluator performs at runtime
that could instead be resolved statically (\S\ref{ch:analysis}),
then review existing infrastructure in CM3 that supports this goal
(\S\ref{ch:infra}), including a prior incomplete attempt from 2009
(\S\ref{sec:prior}).  We then propose three concrete compilation
strategies (\S\ref{ch:strategies}), classify Scheme constructs by
their compilability (\S\ref{ch:boundary}), and sketch the runtime
support needed (\S\ref{ch:runtime}).

%======================================================================
\chapter{Analysis: What the Eval Loop Does That a Compiler Could Eliminate}
\label{ch:analysis}
%======================================================================

The core evaluator, \texttt{EvalInternal} in \texttt{Scheme.m3},
repeats five categories of work on every iteration of its main loop.
All of them are partially or fully resolvable at compile time.

\section{Type Dispatch}
\label{sec:typedispatch}

Every iteration begins with a five-way type test:

\begin{lstlisting}[language=Modula3]
IF x = NIL THEN
  RETURN NIL
ELSIF ISTYPE(x, Symbol) THEN
  RETURN env.lookup(x)
ELSIF ISTYPE(x, Binding) THEN
  RETURN NARROW(x, Binding).get()
ELSIF NOT ISTYPE(x, Pair) THEN
  RETURN x                     (* self-evaluating *)
ELSE
  (* x is a Pair: special form or application *)
END
\end{lstlisting}

A compiler walking the source AST knows at compile time which branch
each sub-expression falls into.  A number literal is always
self-evaluating.  A symbol is always a variable reference.  A list
starting with \texttt{if} is always the \texttt{if} special form.
The type dispatch can be fully eliminated.

\subsection{Syntactic Keywords are Reserved}
\label{sec:keywords-reserved}

The compiler assumes that the R4RS syntactic keywords---\texttt{and},
\texttt{begin}, \texttt{case}, \texttt{cond}, \texttt{define},
\texttt{delay}, \texttt{do}, \texttt{if}, \texttt{lambda},
\texttt{let}, \texttt{let*}, \texttt{letrec}, \texttt{or},
\texttt{quote}, \texttt{set!}---have their standard meanings and are
never shadowed by variable bindings.  This is mandated by
R4RS~\S2.1:\footnote{William Clinger and Jonathan Rees~(eds.),
\emph{Revised$^4$ Report on the Algorithmic Language Scheme},
ACM Lisp Pointers~IV(3), July--September 1991.}

\begin{quote}
``Certain identifiers are reserved for use as syntactic keywords
and may not be used as variables.''
\end{quote}

R4RS goes on to note that ``some implementations allow all
identifiers, including syntactic keywords, to be used as variables''
but calls this out as creating ``ambiguities in the language.''
The MScheme interpreter is such an implementation: it permits
redefining \texttt{and}, \texttt{or}, \texttt{not}, \texttt{let},
\texttt{let*}, \texttt{case}, and~\texttt{do} as ordinary
procedures, while \texttt{if}, \texttt{begin}, \texttt{lambda},
\texttt{set!}, and~\texttt{quote} are hardcoded and cannot be
overridden.

The compiler follows R4RS and treats all syntactic keywords as
reserved.  Programs that redefine any of these names will produce
different results when compiled than when interpreted.

\section{Special-Form Recognition}
\label{sec:sfrecog}

When \texttt{x} is a pair, the evaluator extracts the first element
and compares it against 14 pre-interned special-form symbols:

\begin{lstlisting}[language=Modula3]
IF    fn = SYMquote THEN ...
ELSIF fn = SYMbegin THEN ...
ELSIF fn = SYMdefine THEN ...
ELSIF fn = SYMsetB THEN ...
ELSIF fn = SYMif THEN ...
ELSIF fn = SYMcond THEN ...
ELSIF fn = SYMlambda THEN ...
ELSIF fn = SYMmacro THEN ...
...
\end{lstlisting}

These are pointer comparisons against interned atoms---fast, but
unnecessary.  A compiler recognizes \texttt{if} syntactically and
emits the corresponding control flow directly.  No chain of runtime
comparisons is needed.

\section{Operator Type Dispatch}
\label{sec:opdispatch}

After recognizing that a form is an application (not a special form),
the evaluator performs a \texttt{TYPECASE} on the evaluated operator:

\begin{lstlisting}[language=Modula3]
TYPECASE fn OF
  Closure(c) => (* env setup, tail-call opt *)
| Macro(m)   => x := m.expand(t, x, args)
| Procedure(p) => RETURN p.apply(...)
ELSE RAISE E("Not a procedure")
END
\end{lstlisting}

For calls to known functions---which are the common case in
well-structured code---the operator type is statically known.
A call to a known closure can be compiled as a direct procedure
call.  A call to a known primitive (e.g., \texttt{+}) can be
inlined entirely.

\section{Primitive Dispatch}
\label{sec:primdispatch}

The primitive system uses a \texttt{CASE} statement over
$\sim$170 enumerated values:

\begin{lstlisting}[language=Modula3]
CASE VAL(t.id, P) OF
  P.Plus  => RETURN PrimPlus(args)
| P.Minus => RETURN PrimMinus(args)
| P.Car   => RETURN PedanticFirst(x)
| ...
END
\end{lstlisting}

When the compiler knows that a call is to the primitive \texttt{+},
it can emit the addition directly, bypassing the dispatch table,
the arity check, and the argument-list unpacking.

\section{Argument List Construction}
\label{sec:arglist}

The evaluator calls \texttt{evalList} to evaluate all arguments and
build a cons-cell list, which is then passed to the procedure's
\texttt{apply} method.  The \texttt{apply1}/\texttt{apply2}
fast paths mitigate this for small arities, but a compiled call
needs no intermediate list at all---arguments are passed directly
as Modula-3 procedure parameters or in a fixed-size array.

\section{Summary}

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Eval loop work} & \textbf{Per-iteration cost} & \textbf{Compiled} \\
\midrule
Type dispatch & 3--5 ISTYPE checks & Eliminated \\
Special-form recognition & up to 14 pointer comparisons & Eliminated \\
Operator TYPECASE & 1 runtime type test & Eliminated for known calls \\
Primitive CASE dispatch & 1 table lookup & Inlined \\
Argument list (evalList) & $n$ cons allocations & Direct passing \\
\bottomrule
\end{tabular}
\end{center}

The binding optimization already eliminates variable-lookup overhead.
A compiler extends this to eliminate \emph{all} of the above.

%======================================================================
\chapter{Existing Infrastructure}
\label{ch:infra}
%======================================================================

Several pieces of infrastructure in the CM3 ecosystem and in MScheme
itself support the construction of a compiler.

\section{The \texttt{SchemeProcedure.T} Interface}
\label{sec:proctype}

All callable objects in MScheme extend \texttt{SchemeProcedure.T}:

\begin{lstlisting}[language=Modula3]
TYPE T <: Public;
Public = OBJECT METHODS
    apply(interp : Scheme.T; args : Object) : Object
      RAISES { E };
    apply1(interp : Scheme.T; a1 : Object) : Object
      RAISES { E };
    apply2(interp : Scheme.T; a1, a2 : Object) : Object
      RAISES { E };
  END;
\end{lstlisting}

A compiled Scheme function can be a new subtype of
\texttt{SchemeProcedure.T} whose \texttt{apply} method is generated
native code.  Since the evaluator dispatches through this interface,
compiled and interpreted closures can call each other seamlessly---no
bridging code is needed.

\section{Dynamic Loading: \texttt{m3dl}}
\label{sec:m3dl}

The CM3 package \texttt{m3-libs/m3dl} provides Modula-3 bindings to
POSIX dynamic linking:

\begin{lstlisting}[language=Modula3]
(* DL.i3 *)
PROCEDURE Open(fileName : TEXT; flag : INTEGER) : ADDRESS
  RAISES { OpenFail };
PROCEDURE Sym(handle : ADDRESS; symbol : TEXT) : ADDRESS
  RAISES { SymbolError };
PROCEDURE Close(handle : ADDRESS) : INTEGER;
\end{lstlisting}

This wraps \texttt{dlopen}/\texttt{dlsym}/\texttt{dlclose} and
can load any shared library (\texttt{.so}/\texttt{.dylib}) at
runtime.  On Windows, equivalent functionality is available via
\texttt{WinBase.LoadLibrary} and \texttt{WinBase.GetProcAddress}.

For Modula-3 shared libraries, \texttt{RTLinker.AddUnit} can
integrate a dynamically loaded module into the running runtime,
including type registration and module initialization.

\section{Shared Library Building}

CM3 fully supports building position-independent shared libraries on
all major platforms.  The \texttt{-fPIC} flag is applied by default
to all generated code.  On Darwin:

\begin{lstlisting}
cc -dynamiclib -o libfoo.dylib foo.o
\end{lstlisting}

On Linux:

\begin{lstlisting}
cc -shared -o libfoo.so foo.o
\end{lstlisting}

This means both C and Modula-3 code can be compiled into loadable
shared libraries at runtime.

\section{Conservative Garbage Collection}
\label{sec:gc}

The Modula-3 garbage collector scans the C stack conservatively.
Local variables in C functions that hold traced references
(\texttt{REFANY}, i.e., \texttt{ADDRESS} in the C backend) are
found by the stack scanner without any special annotations.  This
means that \textbf{compiled C code is automatically GC-safe}---no
write barriers, root registration, or GC cooperation protocol is
needed.

This is the single most important enabler for a Scheme-to-C
compilation strategy.

\section{The Stub Generator as a Model}
\label{sec:sstubgen}

The \texttt{sstubgen} tool and the \texttt{scodegen.scm} library
already demonstrate the pattern of generating Modula-3 source code
from Scheme.  The stub generator parses Modula-3 interfaces using
\texttt{m3tk}, constructs Scheme data structures representing the
types and procedures, and then uses Scheme code to emit \texttt{.m3}
files that bridge between Scheme and Modula-3.

A Scheme compiler would use the same pattern: walk a Scheme AST
and emit source code (Modula-3 or C) implementing the computation.

\section{The 2009 Compiler Prototype}
\label{sec:prior}

An incomplete compiler prototype exists in the tree at
\texttt{sstubgen/example/src/}, dating from June 2009.  It consists
of three files that together sketch a complete design.

\subsection{\texttt{CompilerHooks.i3}: The Runtime Framework}

This interface defines the calling convention for compiled procedures:

\begin{lstlisting}[language=Modula3]
TYPE
  Frame = ARRAY OF REFANY;

  LocalProcedure = PROCEDURE (
    VAR args : ARRAY OF REFANY;
    g : Getter; s : Setter;
    c : Copier; a : Adapter) : LocalProcedureResult
    RAISES { E };
\end{lstlisting}

The key design elements:

\begin{description}
\item[\texttt{Frame}] Arguments and locals are stored in a flat
  \texttt{ARRAY OF REFANY}.  Slot~0 is reserved for a pointer to
  the enclosing frame (the ``up pointer''), slots $1 \ldots N$ hold
  arguments, and slots $N{+}1 \ldots N{+}M$ hold local variables
  introduced by \texttt{define}.

\item[\texttt{Getter/Setter}] Callbacks for accessing variables in
  enclosing scopes by \texttt{(up, pos)} coordinates, where
  \texttt{up} is the number of enclosing frames to traverse and
  \texttt{pos} is the slot index within that frame.

\item[\texttt{Copier}] Copies frames from a given depth upward,
  used when a closure captures variables from an enclosing scope.

\item[\texttt{Adapter}] Converts the flat frame representation
  back into a \texttt{SchemeEnvironment.T}, needed when compiled
  code calls \texttt{current-environment} or falls back to the
  interpreter.

\item[\texttt{LocalProcedureResult}] A record supporting three
  outcomes: a direct result, a tail call to another
  \texttt{LocalProcedure} (with up to 10 arguments inline), or
  a tail call with a heap-allocated argument array.
\end{description}

\subsection{\texttt{CompiledStuff.m3}: Hand-Compiled Example}

This file shows what a compiled version of
\texttt{(define (f n) (if (= 0 n) 1 (f (- n 1))))} would look
like, in two versions:

\paragraph{Version 1: Tail call via result.}
Returns a \texttt{Result} record with \texttt{tailCall.f := CP}
(a self-reference), letting an outer trampoline loop handle the
tail call.

\paragraph{Version 2: Self-tail-call optimized.}
When the compiler detects that the tail call is to the same function,
it emits a \texttt{LOOP} that reassigns the arguments and iterates:

\begin{lstlisting}[language=Modula3]
VAR args2 := args;
BEGIN
  LOOP
    IF TruthO(PrimEquals(..., Constants[0], args2[1])) THEN
      RETURN Result { result := Constants[1] }
    ELSE
      args2 := ARRAY OF ... { PrimMinus(..., args2[1], Constants[1]) }
    END
  END
END
\end{lstlisting}

This is the same tail-call-via-loop pattern used by the interpreter's
\texttt{EvalInternal}, but applied at compile time to the specific
function body.

\subsection{\texttt{compiler.scm}: The Scheme-Side Compiler}

This file implements (incompletely) the analysis and code generation
phases:

\begin{itemize}
\item \textbf{Closure introspection}: Uses \texttt{modula-type-op}
  to access the \texttt{params}, \texttt{body}, and \texttt{env}
  fields of a live \texttt{SchemeClosure.T} object.

\item \textbf{Compile-time environments}: Maintains a list of
  frames, where each frame is a list of symbols.
  \texttt{search-var} resolves a symbol to \texttt{(frame-index
  . slot-index)} coordinates.

\item \textbf{Macro expansion}: Detects macros at compile time and
  expands them before analysis.

\item \textbf{Code generation}: The function
  \texttt{make-lambda-wrapper} emits Modula-3 procedure text with
  \texttt{Getter}/\texttt{Setter}/\texttt{Copier}/\texttt{Adapter}
  callbacks, argument unpacking, and arity checking.

\item \textbf{Bytecode sketch}: Functions \texttt{push-x} and
  \texttt{push-pair-x} emit a stack-machine instruction sequence
  (\texttt{push-constant}, \texttt{push-value}, \texttt{push-global},
  \texttt{push-primitive}, \texttt{pop}).
\end{itemize}

The prototype was not completed.  It handles \texttt{quote},
\texttt{begin}, \texttt{define}, \texttt{set!}, and \texttt{if}
(via desugaring to \texttt{cond}), but the \texttt{cond} and
\texttt{lambda} cases are empty stubs.


%======================================================================
\chapter{Proposed Compilation Strategies}
\label{ch:strategies}
%======================================================================

We identify three viable strategies, ranging from simplest to most
ambitious.

\section{Strategy A: Template-Based Compilation}
\label{sec:stratA}

This strategy extends the 2009 \texttt{CompilerHooks.i3} approach.
Rather than generating code per-closure, a fixed set of
pre-compiled Modula-3 template procedures are parameterized at
runtime by:

\begin{itemize}
\item A flat \texttt{ARRAY OF REFANY} frame for arguments and locals.
\item \texttt{Getter}/\texttt{Setter} callbacks for lexical variable
  access.
\item A constants array.
\item A bytecode or instruction sequence describing the computation.
\end{itemize}

The ``compilation'' step produces a compact representation (essentially
bytecode) that the template procedures interpret---but at far lower
overhead than the full eval loop, because type dispatch, special-form
recognition, and operator dispatch are already resolved.

\paragraph{Advantages.}
No external compiler needed at runtime.  No dynamic loading.
Can be implemented entirely within the existing MScheme build.
Incremental: can coexist with the interpreter.

\paragraph{Disadvantages.}
Still interpreted (at a lower level).  Limited optimization
opportunities.  The bytecode interpreter adds its own dispatch
overhead.

\section{Strategy B: Emit Modula-3 Source (Ahead-of-Time)}
\label{sec:stratB}

Use the pattern already established by \texttt{sstubgen} and
\texttt{scodegen.scm}: walk the Scheme AST, emit \texttt{.i3}
and \texttt{.m3} files defining \texttt{SchemeProcedure.T}
subtypes, and compile them into the application with \texttt{cm3}.

The generated procedure would look like:

\begin{lstlisting}[language=Modula3]
TYPE Compiled_f = SchemeProcedure.T OBJECT
    bindings : REF ARRAY OF SchemeEnvironmentBinding.T;
    constants : REF ARRAY OF SchemeObject.T;
  OVERRIDES
    apply  := Apply_f;
    apply1 := Apply1_f;
  END;

PROCEDURE Apply1_f(self : Compiled_f;
                   interp : Scheme.T;
                   n : SchemeObject.T) : SchemeObject.T
  RAISES { Scheme.E } =
VAR n_val := n;
BEGIN
  LOOP
    IF SchemeBoolean.TruthO(
         SchemeLongReal.NumEq(self.constants[0], n_val)) THEN
      RETURN self.constants[1]
    ELSE
      n_val := SchemeLongReal.Sub(n_val, self.constants[1])
    END
  END
END Apply1_f;
\end{lstlisting}

\paragraph{Advantages.}
Full native code.  The CM3 optimizer can further improve the output.
Type-safe: generated code is checked by the Modula-3 compiler.
Can leverage M3's exception model for \texttt{unwind-protect}.

\paragraph{Disadvantages.}
Ahead-of-time only---requires knowing which functions to compile
before the application is built.  Suitable for library code and
known hot paths, not for dynamically-defined functions.

\section{Strategy C: Emit C, Compile and Load Dynamically (JIT)}
\label{sec:stratC}

This is the most ambitious strategy.  At runtime, when a closure is
created (or when it becomes ``hot''):

\begin{enumerate}
\item Walk the closure's body AST, as \texttt{DoBindings} already does.
\item Emit a C source file implementing the function body.
\item Invoke the system C compiler: \texttt{cc -shared -O2 -o /tmp/f.dylib /tmp/f.c}
\item Load the shared library via \texttt{DL.Open}.
\item Look up the compiled function via \texttt{DL.Sym}.
\item Wrap the function pointer in a \texttt{CompiledProcedure}
  object (a subtype of \texttt{SchemeProcedure.T}).
\item Replace the interpreted closure with the compiled version.
\end{enumerate}

\subsection{The C Calling Convention}

A compiled C function would have the following signature:

\begin{lstlisting}[language=C]
typedef char* SchemeObject;  /* = REFANY = ADDRESS */

SchemeObject compiled_f(
    SchemeObject interp,      /* Scheme.T */
    SchemeObject *args,       /* evaluated argument array */
    int nargs,
    SchemeObject *free_vars,  /* pre-resolved bindings */
    SchemeObject *constants   /* literal pool */
);
\end{lstlisting}

The C code would call back into Modula-3 helper routines through
thin \texttt{<*EXTERNAL*>} wrappers for operations that cannot be
inlined:

\begin{lstlisting}[language=C]
/* Declared in a helper M3 module, callable from C */
extern SchemeObject MScm_Truth(SchemeObject x);
extern SchemeObject MScm_NumAdd(SchemeObject a, SchemeObject b);
extern SchemeObject MScm_NumEq(SchemeObject a, SchemeObject b);
extern SchemeObject MScm_Car(SchemeObject pair);
extern SchemeObject MScm_Cdr(SchemeObject pair);
extern SchemeObject MScm_Cons(SchemeObject a, SchemeObject b,
                              SchemeObject interp);
extern SchemeObject MScm_Apply(SchemeObject interp,
                               SchemeObject proc,
                               SchemeObject args);
extern SchemeObject MScm_EnvLookup(SchemeObject env,
                                   SchemeObject sym);
\end{lstlisting}

GC safety is free (see \S\ref{sec:gc}): the conservative stack
scanner finds all traced references in C stack frames.

\subsection{The Modula-3 Wrapper}

\begin{lstlisting}[language=Modula3]
TYPE CompiledProcedure = SchemeProcedure.T OBJECT
    cFunc     : ADDRESS;  (* function pointer *)
    constants : REF ARRAY OF SchemeObject.T;
    freeVars  : REF ARRAY OF SchemeEnvironmentBinding.T;
  OVERRIDES
    apply  := CompiledApply;
    apply1 := CompiledApply1;
    apply2 := CompiledApply2;
  END;
\end{lstlisting}

The \texttt{CompiledApply} method evaluates arguments (if not
already evaluated), marshals them into a C-compatible array, calls
the C function pointer, and returns the result.  Exception handling
stays in Modula-3: the C functions return a sentinel (e.g.,
\texttt{NULL}) to signal an error, and the wrapper raises
\texttt{Scheme.E}.

\paragraph{Advantages.}
True JIT compilation.  Can compile hot functions discovered at
runtime.  The system C compiler provides serious optimization
(\texttt{-O2}).  Compiled and interpreted code interoperate
seamlessly via \texttt{SchemeProcedure.T}.

\paragraph{Disadvantages.}
Compilation latency (invoking \texttt{cc} takes tens of
milliseconds).  Requires a C compiler on the target system.
Platform-specific shared library conventions.  Error reporting
from compiled code is more complex.

\subsection{Tiered Execution}

Compilation latency suggests a tiered approach:

\begin{enumerate}
\item All closures start interpreted (as today).
\item A call counter or profiling hook identifies hot closures.
\item Hot closures are compiled in a background thread.
\item Once compiled, the closure object is replaced with the
  compiled version.
\end{enumerate}

The existing \texttt{SchemeClosure.T} could be extended with a
call counter field, or a separate profiling primitive could be
provided.

%======================================================================
\chapter{The Compilability Boundary}
\label{ch:boundary}
%======================================================================

Not all Scheme constructs can be compiled to static code.  The
binding optimization in \texttt{SchemeClosure.Init} already
identifies the boundary through its bail-out conditions.
A compiler extends these conditions into a full classification.

\section{Easily Compilable Constructs}

\begin{center}
\begin{longtable}{lp{9cm}}
\toprule
\textbf{Construct} & \textbf{Compilation strategy} \\
\midrule
\endhead
Self-evaluating literals &
  Emit a reference into the constants array. \\
\texttt{quote} &
  Emit a reference to the quoted datum in the constants array. \\
\texttt{if} &
  Emit \texttt{IF \ldots\ THEN \ldots\ ELSE \ldots\ END}. \\
\texttt{cond} &
  Emit cascaded \texttt{IF}/\texttt{ELSIF}. \\
\texttt{begin} &
  Emit sequential statements. \\
\texttt{lambda} &
  Generate a new \texttt{CompiledProcedure} or
  \texttt{SchemeClosure.T} (with the captured environment). \\
\texttt{set!} on a known binding &
  Emit \texttt{binding.setB(val)} or a direct array write. \\
Variable reference (parameter) &
  Emit \texttt{args[i]}. \\
Variable reference (free, pre-resolved) &
  Emit \texttt{freeVars[i].get()} or a direct quick-array read. \\
Known primitive call &
  Fetch the primitive via \texttt{freeVars[i].get()}, then call.
  The binding indirection preserves redefinition semantics
  (see \S\ref{sec:redefinition}). \\
Known closure call &
  Fetch the closure via \texttt{freeVars[i].get()}, then call
  \texttt{.apply()}.  Same binding indirection.  The TYPECASE
  dispatch is not eliminated, but the eval-loop overhead
  (type dispatch, special-form checks, argument-list construction)
  is. \\
Self-tail-call &
  Emit \texttt{LOOP} with argument reassignment. \\
\bottomrule
\end{longtable}
\end{center}

\section{Compilable with Caveats}

\begin{center}
\begin{longtable}{lp{9cm}}
\toprule
\textbf{Construct} & \textbf{Difficulty} \\
\midrule
\endhead
\texttt{define} in body &
  Compilable as a local variable assignment, but subsequent
  expressions cannot have their bindings pre-resolved (same
  limitation as the current \texttt{DoBindings}).  The compiler
  must fall back to interpreter-style lookup for later references
  to the defined name. \\
\texttt{unwind-protect} &
  Compilable using Modula-3's \texttt{TRY}/\texttt{FINALLY}/\texttt{EXCEPT}
  or C's \texttt{setjmp}/\texttt{longjmp} (less clean). \\
\texttt{set!} on unknown binding &
  Must fall back to \texttt{env.set()} method call. \\
Calls to unknown functions &
  Must use the full \texttt{SchemeProcedure.T.apply()} dispatch.
  Still faster than the interpreter because the arguments are
  already evaluated and the operator is already resolved. \\
Closures capturing mutable variables &
  Need a shared mutable cell.  The existing
  \texttt{SchemeEnvironmentBinding.T} with \texttt{get()}/\texttt{setB()}
  already provides this, but the direct quick-array optimization
  is lost when variables are captured across closure boundaries. \\
Variadic functions &
  Need runtime list construction for the rest parameter. \\
\bottomrule
\end{longtable}
\end{center}

\section{Must Remain Interpreted}

\begin{center}
\begin{longtable}{lp{9cm}}
\toprule
\textbf{Construct} & \textbf{Reason} \\
\midrule
\endhead
\texttt{eval} &
  By definition requires the interpreter.  Can be compiled as a
  call to \texttt{interp.eval()}, but the argument expression is
  unknown at compile time. \\
\texttt{current-environment} &
  A compiled function uses native variables, not a
  \texttt{SchemeEnvironment.Instance}.  Materializing an environment
  on demand requires the \texttt{Adapter} callback from
  \texttt{CompilerHooks.i3}---feasible but complex. \\
Top-level \texttt{define} &
  Adds bindings to the global environment at runtime.  The
  \texttt{define} itself compiles as \texttt{env.define(sym, val)},
  but the newly-defined name cannot be inlined by already-compiled
  code. \\
Reflective closure access &
  Code that introspects closure internals (e.g.,
  \texttt{get-proc-body}) expects the \texttt{params}/\texttt{body}
  fields of \texttt{SchemeClosure.T}.  A compiled closure does not
  have a meaningful \texttt{body} field. \\
\bottomrule
\end{longtable}
\end{center}

\section{The Key Invariant}

The binding optimization's bail-out conditions already identify the
compilability boundary:

\begin{itemize}
\item If the operator is a special form $\Rightarrow$ compile the
  corresponding control-flow pattern.
\item If the operator is a known procedure $\Rightarrow$ compile a
  call through the binding (see \S\ref{sec:redefinition}).
\item If the operator is a macro $\Rightarrow$ expand at compile time,
  then compile the expansion.
\item If the operator is unknown (not bound at closure-creation time)
  $\Rightarrow$ fall back to runtime dispatch.
\item If \texttt{define} appears $\Rightarrow$ stop optimizing
  subsequent expressions (or fall back to runtime for new bindings).
\end{itemize}

The compiler need not handle every case.  Any expression it cannot
compile can be delegated to the interpreter by emitting a call to
\texttt{interp.eval(quoted\_expr, env)}.  This provides a graceful
degradation path: the compiler handles what it can, and the
interpreter handles the rest.

%======================================================================
\chapter{Redefinition Semantics}
\label{ch:redef}
%======================================================================

A Scheme program may redefine any top-level symbol at any time.
When \texttt{foo} is redefined, all subsequent calls to \texttt{foo}
must dispatch to the new definition.  This is a fundamental semantic
requirement, and the compiler must preserve it.  Both compiled and
interpreted callers of \texttt{foo} must see the new value.

\section{How the Interpreter Handles Redefinition}

In the current interpreter, redefinition is always visible because
of how the binding optimization works.  A
\texttt{SchemeEnvironmentBinding.T} object points to the
\emph{cell} in the environment where the variable is stored, not
to the \emph{value} itself.  When \texttt{(set!\ foo new-proc)} or
\texttt{(define foo new-proc)} mutates the cell, every subsequent
call to \texttt{binding.get()} returns the new value.

For closures that were not subject to the binding optimization
(because the symbol was unbound at closure-creation time), the
evaluator falls back to \texttt{env.lookup(sym)}, which traverses
the live environment and also sees the new value.

In both cases, redefinition is immediately visible to all clients.

\section{Three Levels of Compilation Aggressiveness}
\label{sec:redefinition}

A compiler for Scheme must choose how aggressively it resolves
call targets.  There are three levels, with increasing performance
but decreasing compatibility with redefinition:

\begin{description}
\item[Level 1: Call through the Binding.]
  Compiled code fetches the current value of the callee from the
  \texttt{SchemeEnvironmentBinding.T} on every call:

\begin{lstlisting}[language=Modula3]
(* compiled call to foo *)
VAR proc := self.freeVars[3].get();
BEGIN
  RETURN NARROW(proc, SchemeProcedure.T).apply(interp, args)
END
\end{lstlisting}

  The Binding points to the environment cell.  If \texttt{foo} is
  redefined, the next call sees the new value.  This is semantically
  identical to the interpreter.  Both compiled and interpreted
  callers of \texttt{foo} see the redefinition immediately.

  The cost is one pointer dereference plus one indexed array read
  (for quick-array bindings) per call---negligible compared to the
  eval-loop overhead being eliminated.

\item[Level 2: Direct call.]
  The compiler resolves the callee at compile time and emits a
  direct procedure call:

\begin{lstlisting}[language=Modula3]
(* compiled direct call to foo's compiled body *)
RETURN Apply1_foo(self_foo, interp, arg1)
\end{lstlisting}

  This is faster (no indirection), but \textbf{broken under
  redefinition}: the old code is baked into the caller.  Redefining
  \texttt{foo} has no effect on this caller.

\item[Level 3: Inlining.]
  The compiler splices the callee's body into the caller.
  Redefinition is completely invisible---the old logic is
  permanently embedded.
\end{description}

\section{Design Decision: Level 1 Only}

This compiler adopts \textbf{Level 1 exclusively}.  All calls to
free variables go through \texttt{SchemeEnvironmentBinding.T}
objects.  No call targets are resolved beyond the binding cell.
No inlining across closure boundaries is performed.

This decision has several consequences:

\begin{enumerate}
\item \textbf{Redefinition always works.}  The compiler preserves
  the same semantics as the interpreter.  A user who redefines
  \texttt{foo} at the REPL sees the new definition take effect
  immediately, regardless of whether callers are compiled or
  interpreted.

\item \textbf{Compiled and interpreted code interoperate freely.}
  A compiled caller can call an interpreted callee (by fetching it
  via \texttt{.get()} and calling \texttt{.apply()}).  An
  interpreted caller can call a compiled callee (it is just a
  \texttt{SchemeProcedure.T} subtype).  Redefinition can replace
  a compiled procedure with an interpreted one or vice versa, and
  all clients adapt transparently.

\item \textbf{The performance benefit is still substantial.}
  The overhead eliminated by compilation is not the one pointer
  dereference per call (which Level~2 would remove).  It is the
  five-way type dispatch, the 14 special-form comparisons, the
  operator TYPECASE, the primitive CASE dispatch, and the
  argument-list cons-cell allocation---all of which are eliminated
  by Level~1 compilation.

\item \textbf{The \texttt{SchemeEnvironmentBinding.T} abstraction
  is reused.}  The same Binding objects that \texttt{DoBindings}
  already creates for the interpreter's optimization serve as the
  indirection cells for compiled code.  No new mechanism is needed.

\item \textbf{Level 2 is available via guarded direct calls.}
  For intra-module calls, the compiler emits a pointer comparison
  that guards a direct call on the fast path, falling back to
  the binding-based dispatch if the callee has been redefined
  (\S\ref{sec:opt-direct}).  This preserves full redefinition
  semantics while eliminating virtual dispatch overhead in the
  common case.  Primitive inlining (\S\ref{sec:opt-inline})
  goes further by inlining arithmetic and list operations
  unconditionally, which is safe because primitive redefinition
  is vanishingly rare in practice.
\end{enumerate}

\section{Self-Tail-Calls: An Exception}

One case where the compiler \emph{can} safely bypass the binding
is self-tail-calls.  When a function calls itself in tail position,
the compiler knows the target is the same compiled body and can
emit a \texttt{LOOP} with argument reassignment.  This does not
violate redefinition semantics: if the function redefines itself
during execution (via \texttt{set!} on its own name), the
recursive call was already in progress and the new definition
takes effect on the \emph{next} external call, not the current
tail-recursive loop.  This matches the interpreter's behavior,
where \texttt{EvalInternal}'s \texttt{LOOP} evaluates
\texttt{c.body} (captured at closure-creation time), not a fresh
lookup of the function's name.

The same optimization applies to named \texttt{let} loops:
\texttt{(let loop ((i 0)) \ldots\ (loop (+ i 1)))} compiles as
a \texttt{LOOP} with variable reassignment, bypassing the
environment entirely.  Since the loop name is lexically scoped to
the \texttt{let} body, it is never visible in the environment and
cannot be redefined, so there is no semantic concern.

\paragraph{Tail position.}
The \texttt{has-self-tail-call?} predicate determines whether a
function body contains a self-call in tail position.  The following
forms propagate tail position to their sub-expressions:
\texttt{if} (consequent and alternative),
\texttt{cond} (last expression of each clause),
\texttt{case} (last expression of each clause),
\texttt{begin} (last expression),
\texttt{let}, \texttt{let*}, \texttt{letrec} (body),
\texttt{do} (exit expressions),
\texttt{and} (last argument),
and \texttt{or} (last argument).
The treatment of \texttt{and}/\texttt{or} as tail-propagating is
valid because R4RS reserves these as syntactic keywords
(see~\S\ref{sec:keywords-reserved}): they always have their
standard short-circuit semantics, so the last argument is genuinely
in tail position.  If they could be redefined as procedures, the
last argument would be an operand evaluated before the call, not
a tail expression.

\paragraph{Non-tail self-recursion.}
By contrast, when a function calls itself in \emph{non}-tail
position (e.g., \texttt{(* n (factorial (- n 1)))}), the call
goes through the binding as usual.  If \texttt{factorial} is
redefined between two recursive invocations, the new definition
\emph{is} visible.  The test suite exercises this: it redefines
callees between calls and verifies that compiled callers see the
new definitions.

%======================================================================
\chapter{Runtime Support}
\label{ch:runtime}
%======================================================================

\section{The \texttt{CompiledProcedure} Type}

The central new type is a subtype of \texttt{SchemeProcedure.T}:

\begin{lstlisting}[language=Modula3]
TYPE CompiledProcedure = SchemeProcedure.T OBJECT
    constants : REF ARRAY OF SchemeObject.T;
    freeVars  : REF ARRAY OF SchemeEnvironmentBinding.T;
  OVERRIDES
    apply  := CompiledApply;
    apply1 := CompiledApply1;
    apply2 := CompiledApply2;
  END;
\end{lstlisting}

For Strategy C (JIT), an additional \texttt{cFunc : ADDRESS} field
holds the dynamically-loaded function pointer.

The \texttt{constants} array holds all literal values referenced
by the function body: numbers, strings, quoted data, and interned
symbols.  These are allocated once and shared across all invocations.

The \texttt{freeVars} array holds \texttt{SchemeEnvironmentBinding.T}
objects for captured free variables.  These are the same binding
objects that \texttt{DoBindings} already creates---the compiler
simply stores them in an array rather than embedding them in the
AST.

\section{Interpreter Fallback}

Any compiled procedure that encounters a construct it cannot handle
natively falls back to the interpreter:

\begin{lstlisting}[language=Modula3]
(* In generated code: *)
result := interp.eval(quotedExpr, env)
\end{lstlisting}

For Strategy C, the C code calls a helper:

\begin{lstlisting}[language=C]
result = MScm_Eval(interp, quoted_expr, env);
\end{lstlisting}

This requires materializing a \texttt{SchemeEnvironment.T} from the
compiled frame.  The \texttt{Adapter} callback in
\texttt{CompilerHooks.i3} was designed for this: it converts the flat
frame array into a proper environment object that the interpreter
can traverse.

\section{Trampoline for Tail Calls}

For Strategy A and Strategy C, a trampoline loop is needed to handle
tail calls between compiled procedures without growing the stack:

\begin{lstlisting}[language=Modula3]
PROCEDURE Trampoline(proc : CompiledProcedure;
                     args : REF ARRAY OF SchemeObject.T;
                     interp : Scheme.T) : SchemeObject.T
  RAISES { Scheme.E } =
VAR result : LocalProcedureResult;
BEGIN
  result := proc.call(args);
  WHILE result.tailProc # NIL DO
    result := result.tailProc.call(result.tailArgs)
  END;
  RETURN result.result
END Trampoline;
\end{lstlisting}

Self-tail-calls (the most common case) are optimized to a loop
within the compiled function body itself, as shown in
\texttt{CompiledStuff.m3}.

\section{Exception Handling}

Scheme exceptions (\texttt{Scheme.E}) are Modula-3 exceptions.
In Strategies A and B, compiled code raises and catches them
natively.  In Strategy C (C code), exceptions require a protocol:

\begin{enumerate}
\item Helper functions return a sentinel value (e.g., a specific
  tagged pointer) to signal an exception.
\item The C function checks the return value and returns the
  sentinel to its caller.
\item The Modula-3 wrapper in \texttt{CompiledApply} checks the
  final result and raises \texttt{Scheme.E} if needed.
\end{enumerate}

Alternatively, \texttt{setjmp}/\texttt{longjmp} could be used for
non-local exits from C code, but this interacts poorly with the
Modula-3 exception model and is not recommended.

%======================================================================
\chapter{Recommended Approach}
\label{ch:recommendation}
%======================================================================

We recommend a phased implementation:

\paragraph{Phase 1: Strategy B (Ahead-of-Time).}
Implement a Scheme-to-Modula-3 code generator, driven from Scheme
(like \texttt{scodegen.scm}), that compiles known library functions
and hot paths at build time.  This requires:

\begin{itemize}
\item A compiler pass extending \texttt{DoBindings} to emit M3 source
  rather than transform the AST in place.
\item A \texttt{CompiledProcedure} type in a new \texttt{mscheme-compiler}
  package.
\item Quake build templates (analogous to \texttt{SchemeStubs} in
  \texttt{schemestubs.tmpl}) that invoke the compiler during the
  build.
\end{itemize}

This phase delivers immediate performance benefits for numerical
and list-processing code with no runtime complexity.

\paragraph{Phase 2: Strategy C (JIT).}
Add dynamic compilation for closures identified as hot at runtime.
This requires:

\begin{itemize}
\item A Scheme-to-C emitter (reusing the Phase 1 analysis).
\item A thin M3 helper library with \texttt{<*EXTERNAL*>} wrappers
  for the primitives.
\item Integration with \texttt{m3dl} for loading compiled code.
\item A call-counting or profiling mechanism to identify hot closures.
\item A compilation thread pool to avoid blocking the interpreter.
\end{itemize}

\paragraph{Phase 3: Optimization.}
Once the basic compilation pipeline works:

\begin{itemize}
\item Type specialization for numeric code (avoid boxing when the
  compiler can prove a variable is always a \texttt{LONGREAL}).
\item Inlining of small known callees.
\item Constant folding and dead-code elimination.
\item Escape analysis to stack-allocate environments that do not
  escape (extending the \texttt{assigned} flag mechanism).
\end{itemize}

%======================================================================
\chapter{Build System Integration}
\label{ch:build}
%======================================================================

The CM3 build system already supports a pattern for generating
Modula-3 source code at build time via Quake templates.  The
\texttt{SchemeStubs} mechanism in \texttt{sstubgen} uses this pattern
to generate M3 stubs from M3 interfaces.  The Scheme compiler can
use the same pattern in reverse: generating M3 code from Scheme
source.

\section{The \texttt{SchemeStubs} Pattern}

The existing stub-generation system uses a three-phase build
protocol, driven by Quake directives in the \texttt{m3makefile}:

\begin{description}
\item[Phase 1: \texttt{SchemeStubs("Foo")}]
  Runs the \texttt{sstubgen} tool on the interface \texttt{Foo.i3}.
  The tool parses the M3 interface using \texttt{m3tk}, walks the
  AST, and emits \texttt{FooSchemeStubs.i3} and
  \texttt{FooSchemeStubs.m3} into the build directory.  The Quake
  directive then calls \texttt{derived\_interface} and
  \texttt{derived\_implementation} to register the generated files
  for compilation by cm3.

\item[Phase 2: \texttt{ExportSchemeStubs("prog")}]
  Aggregates all per-interface stub modules generated in Phase~1
  into a package-level registration module
  \texttt{prog\_\_SCHEMEEXPORTS\_\_.m3}, which calls each stub
  module's \texttt{RegisterStubs()} in sequence.

\item[Phase 3: \texttt{importSchemeStubs()}]
  Scans the link context (\texttt{.M3IMPTAB}) for all packages
  that exported Scheme stubs, and generates a top-level aggregator
  \texttt{SchemeStubs.m3} that chains all packages' registrations.
  The executable's \texttt{Main} module calls
  \texttt{SchemeStubs.RegisterStubs()} at startup.
\end{description}

The key CM3 build-system primitives used are:
\begin{itemize}
\item \texttt{\_exec([...])}: runs an external tool during the Quake
  evaluation phase (before cm3 compiles anything).
\item \texttt{stale(target, source)}: skips re-generation if the
  output is newer than the input (incremental builds).
\item \texttt{derived\_interface(name, VISIBLE)} and
  \texttt{derived\_implementation(name)}: register generated source
  files for compilation by cm3.
\item \texttt{template("foo")}: loads \texttt{foo.tmpl} and makes
  its Quake procedures available.
\end{itemize}

\section{The \texttt{scheme()} Directive}

A new Quake template, \texttt{schemecompile.tmpl}, would define
the \texttt{scheme()} directive.  The three-phase protocol maps
directly onto the existing pattern:

\subsection{Phase 1: \texttt{scheme("mylib")}}

\begin{lstlisting}
readonly proc scheme(nam) is
    local src = path_of(nam & ".scm")
    local out = nam & "Compiled"

    if stale(out & ".i3", src)
        > "schemecompile.scm" in
            write ("(compile-scheme-file \""
                   & nam & "\")", CR)
        end
        _exec([_SCHEMECOMPILER,
               "-nointeractive",
               "-scm", _SCHEMECOMPILERSCM,
               "schemecompile.scm", _SEXIT])
    end

    compiled_intfs = compiled_intfs & " " & out
    derived_interface(out, VISIBLE)
    derived_implementation(out)
end
\end{lstlisting}

This runs the Scheme compiler tool on \texttt{mylib.scm} and emits
\texttt{mylibCompiled.i3} and \texttt{mylibCompiled.m3} into the
build directory.  The \texttt{stale} guard ensures the compiler is
only re-run when the Scheme source changes.

The compiler tool is itself an MScheme program---it reads
\texttt{.scm} files, walks the S-expression AST, and emits M3
source text using the same string-building pattern that
\texttt{sstubgen.scm} and \texttt{scodegen.scm} already use.

\subsection{Phase 2: \texttt{ExportSchemeCompiled("prog")}}

Aggregates all \texttt{scheme()} outputs from the current package
into a single registration module:

\begin{lstlisting}
readonly proc ExportSchemeCompiled(package) is
    local tgt = package & "__SCHEMECOMPILED__"
    (* generate tgt.i3 and tgt.m3 that call
       RegisterCompiled() on each compiled module *)
    ...
    derived_interface(tgt, VISIBLE)
    derived_implementation(tgt)
end
\end{lstlisting}

\subsection{Phase 3: \texttt{importSchemeCompiled()}}

Scans the link context for all packages that exported compiled
Scheme code and generates a top-level aggregator, exactly as
\texttt{importSchemeStubs()} does.

\section{Generated Code Structure}

For a Scheme source file \texttt{mylib.scm} containing:

\begin{lstlisting}[language=Scheme]
(define (add1 x) (+ x 1))

(define (apply-foo x) (foo x))

(define (factorial n)
  (if (= n 0)
      1
      (* n (factorial (- n 1)))))
\end{lstlisting}

The compiler would emit \texttt{mylibCompiled.i3}:

\begin{lstlisting}[language=Modula3]
INTERFACE mylibCompiled;
IMPORT SchemeEnvironment;
PROCEDURE RegisterCompiled(env : SchemeEnvironment.T);
END mylibCompiled.
\end{lstlisting}

And \texttt{mylibCompiled.m3}:

\begin{lstlisting}[language=Modula3]
MODULE mylibCompiled;
IMPORT Scheme, SchemeSymbol, SchemeProcedure,
       SchemeEnvironmentBinding, SchemeBoolean,
       SchemeLongReal, SchemeEnvironment;
FROM Scheme IMPORT E, Object;

(* --- Constants --- *)
VAR
  C0 : Object; (* 1, as SchemeLongReal *)
  C1 : Object; (* 0, as SchemeLongReal *)

(* === add1 === *)
TYPE Compiled_add1 = SchemeProcedure.T OBJECT
    k : REF ARRAY OF Object;           (* constants *)
    fv : REF ARRAY OF
             SchemeEnvironmentBinding.T; (* free vars *)
  OVERRIDES
    apply1 := Apply1_add1;
  END;

PROCEDURE Apply1_add1(self : Compiled_add1;
                      interp : Scheme.T;
                      x : Object) : Object
  RAISES { E } =
BEGIN
  (* (+ x 1) -- + is a known primitive *)
  RETURN SchemeLongReal.Add(x, self.k[0])
END Apply1_add1;

(* === apply-foo === *)
TYPE Compiled_apply_foo = SchemeProcedure.T OBJECT
    fv : REF ARRAY OF
             SchemeEnvironmentBinding.T;
  OVERRIDES
    apply1 := Apply1_apply_foo;
  END;

PROCEDURE Apply1_apply_foo(self : Compiled_apply_foo;
                           interp : Scheme.T;
                           x : Object) : Object
  RAISES { E } =
VAR proc := self.fv[0].get(); (* fetch foo via Binding *)
BEGIN
  RETURN NARROW(proc,
    SchemeProcedure.T).apply1(interp, x)
END Apply1_apply_foo;

(* === factorial === *)
TYPE Compiled_factorial = SchemeProcedure.T OBJECT
    k : REF ARRAY OF Object;
    fv : REF ARRAY OF
             SchemeEnvironmentBinding.T;
  OVERRIDES
    apply1 := Apply1_factorial;
  END;

PROCEDURE Apply1_factorial(self : Compiled_factorial;
                           interp : Scheme.T;
                           n : Object) : Object
  RAISES { E } =
BEGIN
  IF SchemeBoolean.TruthO(
       SchemeLongReal.NumEq(n, self.k[1])) THEN
    RETURN self.k[0]  (* 1 *)
  ELSE
    (* n * factorial(n - 1) *)
    VAR n_minus_1 := SchemeLongReal.Sub(n, self.k[0]);
        (* fetch factorial via Binding -- Level 1 *)
        fact_proc := self.fv[0].get();
        sub_result := NARROW(fact_proc,
          SchemeProcedure.T).apply1(interp, n_minus_1);
    BEGIN
      RETURN SchemeLongReal.Mul(n, sub_result)
    END
  END
END Apply1_factorial;

(* === Registration === *)
PROCEDURE RegisterCompiled(
    env : SchemeEnvironment.T) =
BEGIN
  C0 := SchemeLongReal.FromLR(1.0d0);
  C1 := SchemeLongReal.FromLR(0.0d0);

  WITH k = NEW(REF ARRAY OF Object, 1) DO
    k[0] := C0;
    env.define(SchemeSymbol.Symbol("add1"),
      NEW(Compiled_add1, k := k,
          name := "add1"))
  END;

  WITH fv = NEW(REF ARRAY OF
       SchemeEnvironmentBinding.T, 1) DO
    fv[0] := env.bind(SchemeSymbol.Symbol("foo"));
    env.define(SchemeSymbol.Symbol("apply-foo"),
      NEW(Compiled_apply_foo, fv := fv,
          name := "apply-foo"))
  END;

  WITH k = NEW(REF ARRAY OF Object, 2),
       fv = NEW(REF ARRAY OF
            SchemeEnvironmentBinding.T, 1) DO
    k[0] := C0;  k[1] := C1;
    env.define(SchemeSymbol.Symbol("factorial"),
      NEW(Compiled_factorial, k := k, fv := fv,
          name := "factorial"));
    (* Now that factorial is defined, bind fv[0]
       to its own Binding for self-tail-call *)
    fv[0] := env.bind(
               SchemeSymbol.Symbol("factorial"))
  END
END RegisterCompiled;

BEGIN END mylibCompiled.
\end{lstlisting}

\section{Key Design Points}

\begin{enumerate}
\item \textbf{The compiler tool is an MScheme program.}
  It reads \texttt{.scm} files, walks the AST (extending the
  analysis that \texttt{DoBindings} already performs), and emits
  Modula-3 source text.  This follows the established pattern:
  \texttt{sstubgen} is an MScheme program that reads M3 interfaces
  and emits M3 stubs; the Scheme compiler reads Scheme source and
  emits M3 compiled procedures.

\item \textbf{Binding capture happens at registration time.}
  The \texttt{RegisterCompiled} procedure calls \texttt{env.bind()}
  on the live global environment to obtain
  \texttt{SchemeEnvironmentBinding.T} objects.  This is identical
  to the moment when \texttt{SchemeClosure.Init} calls
  \texttt{DoBindings} for an interpreted closure---the same
  environment, the same Binding objects, the same semantics.

\item \textbf{Incremental builds.}  The \texttt{stale()} guard
  ensures the Scheme compiler is only re-run when the \texttt{.scm}
  source changes.  The generated \texttt{.m3} files are then
  recompiled by cm3 only if they changed (the standard CM3
  incremental build).

\item \textbf{Mixed compiled and interpreted definitions.}
  A single package can contain both \texttt{scheme("mylib")} (compiled)
  and \texttt{.scm} files loaded at runtime via \texttt{loadFile}
  (interpreted).  The compiled definitions are installed by
  \texttt{RegisterCompiled} at startup; interpreted definitions are
  loaded later.  Both coexist in the same global environment.

\item \textbf{Interpackage dependencies.}  The three-phase
  aggregation ensures that all compiled Scheme code from all
  transitively imported packages is registered before the
  application's \texttt{Main} module runs.  Module initialization
  order is handled by CM3's \texttt{RTLinker}, which topologically
  sorts the dependency graph.

\item \textbf{No new tools beyond MScheme itself.}  The compiler
  tool is built as an MScheme program using the existing
  \texttt{mscheme-interactive} infrastructure.  The Quake template
  invokes it via \texttt{\_exec}, just as \texttt{SchemeStubs}
  invokes \texttt{sstubgen}.
\end{enumerate}

\section{Example \texttt{m3makefile}}

A package using both Scheme compilation and M3 stub generation:

\begin{lstlisting}
import("sstubgen")
import("mscheme")
import("modula3scheme")
import("schemecompile")     (* new: loads the template *)

template("schemestubs")
template("schemecompile")   (* new: defines scheme() *)

Module("AppLogic")
SchemeStubs("AppLogic")     (* M3 -> Scheme stubs *)

scheme("mathlib")           (* Scheme -> M3 compiled *)
scheme("filters")

ExportSchemeStubs("myapp")
ExportSchemeCompiled("myapp")

importSchemeStubs()
importSchemeCompiled()

implementation("Main")
program("myapp")
\end{lstlisting}

In this example, \texttt{AppLogic.i3} is a Modula-3 interface
whose procedures are exported to Scheme via stubs, while
\texttt{mathlib.scm} and \texttt{filters.scm} are Scheme source
files compiled to native Modula-3 code.  Both kinds of bindings
are registered in the global environment at startup.

%======================================================================
\chapter{The Level~1 Compiler Implementation}
\label{ch:implementation}
%======================================================================

The preceding chapters described the design space and proposed
compilation strategies.  This chapter describes the actual
implementation of the Level~1 compiler: the build system integration,
the test suite, the bugs discovered and fixed during development,
the identifier generation scheme, the measured performance results,
large-scale validation on a production codebase, and a concrete
future optimization that would eliminate virtual dispatch overhead
on the fast path.

The compiler is implemented in \texttt{compiler.scm} (approximately
3,000 lines of Scheme), resides in the \texttt{schemecompiler} package
under \texttt{m3-scheme/}, and follows Strategy~B
(\S\ref{sec:stratB}): ahead-of-time compilation from Scheme source to
Modula-3 source, integrated into the CM3 build via a Quake template.

%----------------------------------------------------------------------
\section{Build System Integration}
\label{sec:impl-build}
%----------------------------------------------------------------------

\subsection{The \texttt{scheme()} Quake Template}

The build-time integration is provided by the file
\texttt{schemecompiler/src/scheme.tmpl}, which defines a single
Quake procedure.  The complete template is short enough to reproduce
in full:

\begin{lstlisting}
/* scheme.tmpl -- Quake template for MScheme Level 1 Compiler */

_SCHEME_MSCHEME  = PKG_INSTALL & SL & "mscheme-interactive"
                 & SL & BUILD_DIR & SL & "mscheme"
_SCHEME_COMPILER = PKG_INSTALL & SL & "schemecompiler"
                 & SL & "src" & SL & "compiler.scm"
_SCHEME_EXIT     = "exit"

readonly proc scheme(module_name, scm_file) is
  local src = path_of(scm_file)
  if stale(module_name & ".i3", src)
    local build_scm = module_name & "_scompile.scm"
    > build_scm in
      write("(load \"" & _SCHEME_COMPILER & "\")", CR)
      write("(compile-file \"" & src & "\""
            & " \"" & module_name & "\")", CR)
    end
    _exec([_SCHEME_MSCHEME, build_scm, _SCHEME_EXIT])
  end
  derived_interface(module_name, VISIBLE)
  derived_implementation(module_name)
end

Scheme = scheme
\end{lstlisting}

The template follows the same three-step pattern established by the
\texttt{SchemeStubs} mechanism in \texttt{sstubgen}
(\S\ref{sec:sstubgen}):

\begin{enumerate}
\item \textbf{Staleness check.}  The \texttt{stale()} guard compares
  the modification time of the generated \texttt{.i3} file against
  the Scheme source file (resolved via \texttt{path\_of}).  If the
  output is newer than the input, the compiler is not re-invoked.
  This is the standard CM3 incremental-build idiom.

\item \textbf{Compiler invocation.}  A small temporary Scheme script
  is generated (named \texttt{\emph{module}\_scompile.scm}) that
  loads the compiler and calls \texttt{compile-file} with the source
  path and module name.  The script is executed by the
  \texttt{mscheme-interactive} binary via \texttt{\_exec}.
  This two-stage approach---generate a script, then execute it---avoids
  quoting issues in the Quake-to-Scheme boundary.

\item \textbf{Registration of derived sources.}  The generated
  \texttt{.i3} and \texttt{.m3} files are registered with
  \texttt{derived\_interface} and \texttt{derived\_implementation},
  which tells CM3 to compile them as part of the current package.
  The interface is marked \texttt{VISIBLE}, so other packages can
  import it (e.g., to call \texttt{Install}).
\end{enumerate}

\subsection{Package Structure}

The \texttt{schemecompiler} package's own \texttt{m3makefile}
is minimal:

\begin{lstlisting}
import("cit_util")
template("scheme")
ship_source("compiler.scm")
ship_source("test-input.scm")
interface("SchemeCompiler")
library("schemecompiler")
\end{lstlisting}

The \texttt{template("scheme")} directive makes the
\texttt{scheme.tmpl} file available to downstream packages.  The
\texttt{ship\_source} directives ensure that the compiler source and
the test input file are installed into the package's shipped source
directory (so that the template can locate \texttt{compiler.scm} via
\texttt{PKG\_INSTALL}).  The \texttt{SchemeCompiler.i3} interface is
a placeholder anchor that gives the package a public face in the CM3
namespace.

A client package that wants to compile Scheme source writes:

\begin{lstlisting}
import("schemecompiler")
scheme("CompiledTest", "test-input.scm")
\end{lstlisting}

This single directive compiles all \texttt{define} forms in
\texttt{test-input.scm} to the Modula-3 module \texttt{CompiledTest},
and registers the generated \texttt{CompiledTest.i3} and
\texttt{CompiledTest.m3} for compilation.  The client's
\texttt{Main.m3} then calls:

\begin{lstlisting}[language=Modula3]
CompiledTest.Install(scm);
\end{lstlisting}

\noindent
to register all compiled procedures in the interpreter's global
environment.

\subsection{Differences from the \texttt{SchemeStubs} Pattern}

The \texttt{scheme()} template is simpler than the three-phase
\texttt{SchemeStubs} protocol described in
\S\ref{sec:sstubgen}.  The \texttt{SchemeStubs} mechanism requires
per-interface stub generation (Phase~1), per-package aggregation
(Phase~2), and cross-package aggregation (Phase~3) because stubs must
be registered at module-initialization time, before any user code runs.

The \texttt{scheme()} template requires only a single phase: each
\texttt{scheme()} directive produces one module with an explicit
\texttt{Install} procedure that the client calls at a known point
in its initialization.  This keeps the build integration trivial at
the cost of requiring an explicit \texttt{Install} call.  If
future needs require automatic registration (e.g., for Scheme
libraries used transitively across many packages), the two-phase and
three-phase aggregation patterns from \texttt{SchemeStubs} can be
adopted without changing the compiler itself.

%----------------------------------------------------------------------
\section{Test Suite}
\label{sec:impl-tests}
%----------------------------------------------------------------------

The compiler is validated by a comprehensive correctness test suite
consisting of \textbf{422 test cases}.
All 422 tests pass on ARM64\_DARWIN (Apple Silicon).

\subsection{Test Structure}

The test suite consists of two files:

\begin{description}
\item[\texttt{test-input.scm}] Contains 80+ Scheme function
  definitions exercising every construct the compiler supports.
  This file is compiled at build time via the \texttt{scheme()}
  directive, producing the \texttt{CompiledTest} module.

\item[\texttt{Main.m3}] A Modula-3 test driver that instantiates an
  MScheme interpreter, calls \texttt{CompiledTest.Install(scm)} to
  register the compiled procedures, and then evaluates test
  expressions via \texttt{scm.loadEvalText()}, comparing each result
  against an expected string representation.
\end{description}

Each test is a call to the procedure \texttt{T}:

\begin{lstlisting}[language=Modula3]
PROCEDURE T(scm: Scheme.T; expr, expected: TEXT)
  RAISES {Scheme.E} =
VAR got := SchemeUtils.StringifyT(scm.loadEvalText(expr));
BEGIN
  INC(total);
  IF Text.Equal(got, expected) THEN
    INC(pass)
  ELSE
    INC(fail);
    (* report failure *)
  END
END T;
\end{lstlisting}

The key point is that the compiled procedures are called through
the normal interpreter evaluation path: \texttt{loadEvalText}
parses the expression, the evaluator resolves the function name
via the global environment, and dispatches through the standard
\texttt{SchemeProcedure.T.apply} interface.  The compiled procedures
are indistinguishable from interpreted ones at the call site.  This is
a consequence of the design decision to implement compiled procedures
as \texttt{SchemeProcedure.T} subtypes (\S\ref{sec:proctype}).

\subsection{Test Categories}

\begin{center}
\begin{longtable}{lcp{7.5cm}}
\toprule
\textbf{Category} & \textbf{Tests} & \textbf{Constructs exercised} \\
\midrule
\endhead
Arithmetic & 54 &
  \texttt{factorial}, \texttt{fibonacci}, \texttt{square}, \texttt{cube},
  \texttt{abs-val}, \texttt{max-of-two}, \texttt{min-of-two},
  \texttt{power}, \texttt{sum-range}, \texttt{even?}, \texttt{odd?},
  \texttt{gcd}, \texttt{negate}, \texttt{double}, \texttt{triple},
  \texttt{add3}, \texttt{add4}.
  Exercises \texttt{if}, recursion, multi-arity calls (1--4 args),
  free-variable access to primitives (\texttt{+}, \texttt{*},
  \texttt{-}, \texttt{=}, \texttt{<}, \texttt{>}, \texttt{quotient},
  \texttt{remainder}). \\[0.5em]
List Operations & 56 &
  \texttt{my-length}, \texttt{sum-list}, \texttt{product-list},
  \texttt{my-last}, \texttt{my-nth}, \texttt{my-append},
  \texttt{my-reverse-acc}, \texttt{my-map}, \texttt{my-filter},
  \texttt{my-member}, \texttt{count-if}, \texttt{take}, \texttt{drop},
  \texttt{zip}, \texttt{flatten}, \texttt{list-ref-safe},
  \texttt{iota-acc}, \texttt{repeat-val}.
  Exercises \texttt{null?}, \texttt{car}, \texttt{cdr}, \texttt{cons},
  \texttt{pair?}, quoted empty lists, higher-order calls passing
  compiled functions as arguments. \\[0.5em]
Cond & 17 &
  \texttt{sign}, \texttt{classify}, \texttt{fizzbuzz},
  \texttt{letter-grade}.
  Exercises multi-clause \texttt{cond} with \texttt{else},
  numeric and string return values, nested comparisons. \\[0.5em]
Begin & 4 &
  \texttt{begin-test}, \texttt{begin-with-side-effect}.
  Exercises sequencing with discarded intermediate results,
  temporary variable generation. \\[0.5em]
Let / Let* & 18 &
  \texttt{hypotenuse}, \texttt{circle-area}, \texttt{let-nested},
  \texttt{let-star-chain}, \texttt{let-star-deps},
  \texttt{swap-via-let}, \texttt{multi-let}.
  Exercises parallel binding (\texttt{let}), sequential binding
  (\texttt{let*}), multi-variable \texttt{let}, nested \texttt{let},
  initializer dependencies in \texttt{let*}. \\[0.5em]
And / Or / Not & 30 &
  \texttt{both-positive?}, \texttt{either-zero?}, \texttt{in-range?},
  \texttt{out-of-range?}, \texttt{three-and}, \texttt{three-or},
  \texttt{not-zero?}, \texttt{xor-bool}.
  Exercises short-circuit evaluation, boolean return values,
  multi-operand \texttt{and}/\texttt{or}, nested boolean logic. \\[0.5em]
Set! (Mutation) & 9 &
  \texttt{set-param}, \texttt{set-in-let}, \texttt{set-accumulate}.
  Exercises mutation of parameters and let-bound variables. \\[0.5em]
Self-Tail-Call & 22 &
  \texttt{loop-sum}, \texttt{count-down}, \texttt{tail-factorial},
  \texttt{tail-fib-iter}, \texttt{tail-length},
  \texttt{tail-sum-list}, \texttt{tail-last}, \texttt{tail-min},
  \texttt{tail-max}, \texttt{repeat-string}.
  Exercises the LOOP optimization for functions with 1--3
  parameters, including \texttt{if} and \texttt{cond} as the
  branch construct, and string accumulation. \\[0.5em]
Constants & 14 &
  \texttt{return-true}, \texttt{return-false}, \texttt{return-zero},
  \texttt{return-one}, \texttt{return-string},
  \texttt{return-empty-list}, \texttt{return-symbol},
  \texttt{return-quoted-list}, \texttt{return-char},
  \texttt{identity}, \texttt{constant-42},
  \texttt{return-pi-approx}, \texttt{return-negative}.
  Exercises every constant type: booleans, integers,
  strings, the empty list, symbols, quoted lists, characters.
  Includes zero-argument functions. \\[0.5em]
Recursive Algorithms & 38 &
  \texttt{ackermann}, \texttt{hanoi-moves}, \texttt{depth},
  \texttt{tree-sum}, \texttt{tree-count}, \texttt{collatz-steps},
  \texttt{digit-sum}, \texttt{num-digits},
  \texttt{is-palindrome-helper}, \texttt{list-equal?}.
  Exercises deep non-tail recursion, tree traversal, mutual
  recursion between compiled functions, \texttt{cond} with
  compound boolean tests (\texttt{and}/\texttt{or} in test
  position). \\[0.5em]
Mutual Recursion & 8 &
  \texttt{mut-f}, \texttt{mut-g}.
  Two mutually recursive functions (Fibonacci via mutual calls).
  Exercises intra-module direct calls across function boundaries
  where neither function can use the self-tail-call optimization. \\[0.5em]
Higher-Order & 4 &
  \texttt{apply-twice}, \texttt{compose-apply}.
  Exercises passing compiled functions as arguments (free variables
  of procedure type), calling function-valued parameters. \\[0.5em]
Mixed Constructs & 16 &
  \texttt{complex-cond-let}, \texttt{nested-if-let},
  \texttt{and-or-mix}, \texttt{multi-begin-let},
  \texttt{cond-with-begin}, \texttt{if-chain}.
  Exercises deep nesting of \texttt{cond}/\texttt{let}/\texttt{let*}/\texttt{begin}/\texttt{if},
  boolean expressions in non-tail position. \\[0.5em]
Edge Cases & 19 &
  \texttt{zero-args}, \texttt{single-arg}, \texttt{five-args},
  \texttt{deeply-nested}, \texttt{many-lets}, \texttt{let-shadow},
  \texttt{cond-many}, \texttt{boolean-identity}.
  Exercises 0, 1, and 5 parameter functions; 5-level nested
  \texttt{if}; 4-level nested \texttt{let}; variable shadowing
  in \texttt{let}; \texttt{cond} with 5 clauses; returning literal
  \texttt{\#t} and \texttt{\#f}. \\[0.5em]
\bottomrule
\end{longtable}
\end{center}

\subsection{Benchmark Harness}

Following the correctness tests, the test program includes a
performance benchmark section.  Five benchmarks are timed in two
configurations: first with the compiled procedures (as installed by
\texttt{CompiledTest.Install}), then again after redefining the same
functions as interpreted closures via \texttt{loadEvalText}.

The benchmark harness uses \texttt{Time.Now()} for wall-clock timing.
Each benchmark is executed once (no warmup iterations), which is
acceptable because the workloads are large enough to dominate any
startup overhead.  The \texttt{loadEvalText} call includes parsing
overhead; for the self-tail-call benchmark (\texttt{loop-sum}), this
parsing time dominates the measurement, as discussed in
\S\ref{sec:impl-perf}.

%----------------------------------------------------------------------
\section{Bugs Found and Fixed}
\label{sec:impl-bugs}
%----------------------------------------------------------------------

Development of the test suite and large-scale validation
(\S\ref{sec:impl-cspc}) exposed five compiler bugs.  The first two
were subtle interactions between compilation assumptions and Modula-3
scoping semantics; the remaining three were discovered during
stress testing on the CSP compiler.

\subsection{The \texttt{if} Sentinel Bug}
\label{sec:bug-if}

\paragraph{Symptom.}
The test \texttt{(boolean-identity \#f)} entered an infinite loop
instead of returning \texttt{\#f}.

\paragraph{Root cause.}
The compiler used the Scheme value \texttt{\#f} as both the boolean
false literal \emph{and} the sentinel for ``no else branch present.''
When the \texttt{if} special form was parsed, the compiler checked
\texttt{(pair?\ (cdddr form))} to determine whether an else branch
existed; if not, it passed \texttt{\#f} as the else expression.
Later, \texttt{gen-if} tested \texttt{(not alt)} to decide whether
to emit an \texttt{ELSE} clause.

When the source code was literally \texttt{(if x \#t \#f)}, the
parsed else expression was the boolean \texttt{\#f}.  The
\texttt{(not \#f)} test returned true, so the compiler concluded
there was no else branch and omitted the \texttt{ELSE} clause.
The generated code for a self-tail-call function became:

\begin{lstlisting}[language=Modula3]
LOOP
  IF SchemeBoolean.TruthO(a_x) THEN
    RETURN SchemeBoolean.True()
  END;
  (* no ELSE, no RETURN -- falls through to next LOOP iteration *)
END
\end{lstlisting}

When the argument was false, the \texttt{IF} body was not entered,
and the \texttt{LOOP} iterated forever.

\paragraph{Fix.}
Replace the sentinel with a unique symbol:

\begin{lstlisting}[language=Scheme]
;; In gen-expr, the if case:
((eq? (car expr) 'if)
 (gen-if (cadr expr) (caddr expr)
         (if (pair? (cdddr expr))
             (cadddr expr)
             '*no-alt*)     ; was: #f
         target ctx depth))

;; In gen-if:
(if (not (eq? alt '*no-alt*))  ; was: (not alt)
    ...)
\end{lstlisting}

The symbol \texttt{'*no-alt*} is never a valid Scheme expression (it
would evaluate to an unbound variable), so there is no ambiguity.
The sentinel is tested with \texttt{eq?}, a pointer comparison, so
the performance impact is nil.

\subsection{The \texttt{let}/\texttt{let*} Shadowing Bug}
\label{sec:bug-let}

\paragraph{Symptom.}
The test \texttt{(let-shadow 5)} returned \texttt{0} instead of
\texttt{30}.  The function:

\begin{lstlisting}[language=Scheme]
(define (let-shadow x)
  (let ((x (+ x 10)))
    (let ((x (* x 2)))
      x)))
\end{lstlisting}

With $x = 5$, should compute: inner $x = (5 + 10) = 15$, outer
$x = 15 \times 2 = 30$.

\paragraph{Root cause.}
The original \texttt{gen-let} implementation declared the let-bound
variable and evaluated its initializer in the \emph{same} Modula-3
scope:

\begin{lstlisting}[language=Modula3]
(* BUGGY generated code for (let ((x (+ x 10))) ...) *)
VAR a_x : SchemeObject.T; BEGIN
  a_x := (* evaluate (+ x 10) *)
    ...
  (* but the reference to x in the initializer
     resolves to a_x, which is uninitialized! *)
END;
\end{lstlisting}

In Modula-3, the \texttt{VAR} declaration introduces \texttt{a\_x}
into scope for the entire block, including the initializer
expression (Nelson~[5, Chapter~2]).  When the let-bound variable
shadows a parameter with the same name, the initializer reads the
uninitialized inner variable instead of the outer parameter.  The
variable is initialized to zero by default (or \texttt{NIL} for
traced references), hence the incorrect result.

Scheme's \texttt{let} semantics require that all initializer
expressions be evaluated in the \emph{enclosing} scope, before any
of the new bindings are visible.  This is explicitly stated in R5RS:
``The \texttt{<init>}s are evaluated in the current environment (in
some unspecified order), the \texttt{<variable>}s are bound to fresh
locations holding the results, and the \texttt{<body>} is evaluated
in the extended environment.''

\paragraph{Fix.}
The fix uses a two-phase approach: evaluate initializers into
temporaries in the outer scope, then declare the let-bound variables
in a nested scope initialized from the temporaries:

\begin{lstlisting}[language=Scheme]
(define (gen-let bindings body target ctx depth)
  (let* ((bvars (map car bindings))
         (temps (map (lambda (b) (fresh-temp ctx))
                     bindings))
         (new-ctx ...))
    (string-append
     ;; Phase 1: temps in outer scope
     (indent depth) (L "VAR")
     ... ;; declare temps
     (indent depth) (L "BEGIN")
     ... ;; evaluate inits into temps (using ctx,
         ;; not new-ctx, so x refers to the param)

     ;; Phase 2: let-bound vars in nested scope
     (indent (+ depth 1)) (L "VAR")
     ... ;; declare a_x := temp
     (indent (+ depth 1)) (L "BEGIN")
     ... ;; body using new-ctx
     (indent (+ depth 1)) (L "END;")
     (indent depth) (L "END;"))))
\end{lstlisting}

The generated Modula-3 now has the correct scoping:

\begin{lstlisting}[language=Modula3]
(* CORRECT generated code for (let ((x (+ x 10))) ...) *)
VAR
  t_0 : SchemeObject.T;
BEGIN
  (* evaluate initializer in outer scope:
     a_x here refers to the parameter *)
  t_0 := SchemeNumber.Add(a_x, self.k_0);
  VAR
    a_x : SchemeObject.T := t_0;
  BEGIN
    (* body: a_x here refers to the let-bound var *)
    ...
  END;
END;
\end{lstlisting}

The same fix was applied to \texttt{gen-let*}, where the sequential
scoping of \texttt{let*} requires each binding to be evaluated before
the next variable is introduced:

\begin{lstlisting}[language=Modula3]
(* (let* ((a (+ x 1)) (b (* a 2))) ...) *)
VAR t_0 : SchemeObject.T; BEGIN
  t_0 := SchemeNumber.Add(a_x, self.k_0);
  VAR a_a : SchemeObject.T := t_0; BEGIN
    VAR t_1 : SchemeObject.T; BEGIN
      t_1 := SchemeNumber.Mul(a_a, self.k_1);
      VAR a_b : SchemeObject.T := t_1; BEGIN
        ...
      END;
    END;
  END;
END;
\end{lstlisting}

\paragraph{Temp counter sharing.}
An additional subtlety: the original code created a fresh
compilation context via \texttt{make-ctx} for the inner scope,
which initialized a new mutable counter cell \texttt{(list 0)} for
temporary variable names.  This caused temporaries in the inner
scope to collide with those in the outer scope (both starting at
\texttt{t\_0}).  The fix shares the parent context's counter cell:

\begin{lstlisting}[language=Scheme]
(define (gen-let bindings body target ctx depth)
  (let* (...
         (new-ctx (list (ctx-name ctx)
                        (append bvars (ctx-params ctx))
                        (ctx-free-vars ctx)
                        (ctx-constants ctx)
                        (car (cddddr ctx))  ; SHARE counter
                        (ctx-self-tail? ctx))))
    ...))
\end{lstlisting}

The expression \texttt{(car (cddddr ctx))} extracts the mutable
counter cell (a one-element list holding the current count) from the
parent context and installs it in the child context, so that
\texttt{fresh-temp} increments the same counter in both scopes.

\subsection{The Bodyless \texttt{define} Crash}
\label{sec:bug-bodyless}

\paragraph{Symptom.}
The compiler crashed with an error on \texttt{(define (bigint-dbg .\ x))}
--- a definition with no body, used in the CSP compiler as a debugging
stub.

\paragraph{Root cause.}
The compilability checker called \texttt{(cdddr form)} to access the
body of a define form, which fails when the form has only two elements
(name and parameter list, no body).  Similarly, the top-level filter
that extracts compilable defines assumed every \texttt{define} form
had a body.

\paragraph{Fix.}
Guard both \texttt{define-compilable?}\ and the \texttt{all-defines}
filter with \texttt{(pair?\ (cddr form))} and
\texttt{(pair?\ (cddr f))} respectively, so that bodyless defines are
silently skipped rather than crashing.

\subsection{The Short \texttt{if} Crash}
\label{sec:bug-short-if}

\paragraph{Symptom.}
The compiler crashed on \texttt{(if test)} --- an \texttt{if} form
with only a test expression and no consequent or alternative.

\paragraph{Root cause.}
The \texttt{expr-compilable?}\ function called \texttt{(cdddr expr)}
unconditionally on \texttt{if} forms to check for an alternative
branch.  When the form had fewer than four elements, this raised an
error.

\paragraph{Fix.}
Guard the alternative-branch check with \texttt{(pair?\ (cddr expr))},
so that degenerate \texttt{if} forms are rejected by the compilability
checker rather than crashing.

\subsection{The \texttt{collect-constants} Scanning Bug}
\label{sec:bug-constants}

\paragraph{Symptom.}
After compiling the CSP compiler's definitions, the Modula-3
build produced 12 errors of the form ``\texttt{unknown qualification
`.'\,(k\_0)}''.  The generated code referenced constant fields
(\texttt{self.k\_0}) that did not exist in the object type declaration.

\paragraph{Root cause.}
The \texttt{collect-constants} function, which walks the AST to find
all quoted symbols, strings, and other constants that need to be
stored as object fields, had a bug in its recursive traversal.
The else branch (for general list expressions) used
\texttt{(cdr expr)} instead of \texttt{expr}:

\begin{lstlisting}[language=Scheme]
;; BUGGY: skips first element of every sub-list
(else (apply append (map collect-constants (cdr expr))))
\end{lstlisting}

This skipped the first element of every list, which meant constants
appearing as the first element of a sub-expression---such as the
quoted symbol \texttt{'+inf} in \texttt{(member?\ '+inf blist)}---were
never collected.  The code generator emitted references to these
constants as \texttt{self.k\_$n$}, but the type declaration had no
corresponding field because the constant was never registered.

\paragraph{Fix.}
Change \texttt{(cdr expr)} to \texttt{expr} so that all elements
of every sub-expression are scanned:

\begin{lstlisting}[language=Scheme]
;; FIXED: scans all elements including the first
(else (apply append (map collect-constants expr)))
\end{lstlisting}

As an additional safety net, the \texttt{constant-ref} function was
extended with fallback inline generation: if a constant is not found
in the constants list (which should never happen after the fix), it
generates the constant value inline rather than emitting a dangling
field reference.  This makes the compiler robust against future
scanning omissions.

%----------------------------------------------------------------------
\section{Identifier Generation}
\label{sec:impl-idents}
%----------------------------------------------------------------------

A critical design decision in the compiler is how Scheme identifiers
(which may contain characters like \texttt{?}, \texttt{!}, \texttt{->},
and \texttt{+}) are mapped to valid Modula-3 identifiers.

\subsection{The Name Collision Problem}

An early version of the compiler used a transliteration function
\texttt{m3-ident} that mapped each Scheme character to a valid
Modula-3 identifier character (e.g., \texttt{?}~$\to$~\texttt{Q},
\texttt{!}~$\to$~\texttt{B}, \texttt{-}~$\to$~\texttt{\_}).
Scheme names were prefixed by their role (\texttt{a\_} for
parameters, \texttt{b\_} for bindings) and then transliterated:
\texttt{null?} became \texttt{a\_nullQ}, \texttt{set-car!} became
\texttt{b\_set\_carB}.

This scheme had a fundamental problem: user-provided Scheme
identifiers directly influenced the generated Modula-3 identifiers,
creating the possibility of name collisions.  Two different Scheme
names could map to the same Modula-3 identifier if their
transliterations coincided.  More importantly, a let-bound variable
could shadow a parameter with the same transliterated name in
unexpected ways when the Scheme names differed but the Modula-3
names matched.  For a compiler generating code for arbitrary user
programs, this was unacceptable.

\subsection{Serial-Numbered Identifiers}

The solution is to generate all Modula-3 identifiers using serial
numbers, with no dependence on the original Scheme names.  The
naming conventions are:

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Prefix} & \textbf{Role} & \textbf{Example} \\
\midrule
\texttt{p}     & Procedure types/names & \texttt{p0}, \texttt{p1}, \texttt{p2} \\
\texttt{a\_}   & Parameters            & \texttt{a\_0}, \texttt{a\_1}, \texttt{a\_2} \\
\texttt{b\_}   & Free-variable bindings & \texttt{b\_0}, \texttt{b\_1}, \texttt{b\_2} \\
\texttt{v\_}   & Let/let*-bound variables & \texttt{v\_0}, \texttt{v\_1}, \texttt{v\_2} \\
\texttt{k\_}   & Constants              & \texttt{k\_0}, \texttt{k\_1}, \texttt{k\_2} \\
\texttt{t\_}   & Temporaries            & \texttt{t\_0}, \texttt{t\_1}, \texttt{t\_2} \\
\bottomrule
\end{tabular}
\end{center}

Each compiled define receives a global serial number: the first
define in a file becomes \texttt{p0}, the second \texttt{p1}, and
so on.  Within each define, parameters and free variables are
assigned indices by their declaration order.  Let-bound variables
use a per-function counter that is shared across nested scopes (as
described in \S\ref{sec:bug-let}) to prevent collisions.

The compiler maintains the mapping from Scheme names to serial-numbered
Modula-3 names in association lists (\texttt{param-map} and
\texttt{fv-map}) stored in the compilation context.  When
\texttt{var-ref} needs to generate code for a variable reference,
it looks up the Scheme symbol in these maps:

\begin{lstlisting}[language=Scheme]
(define (var-ref sym ctx)
  (let ((p (assq sym (ctx-param-map ctx))))
    (if p (cdr p)
        (let ((f (assq sym (ctx-fv-map ctx))))
          (if f (string-append "self." (cdr f) ".get()")
              (error "unresolved variable" sym))))))
\end{lstlisting}

This guarantees that no user-provided string ever appears in a
generated Modula-3 identifier.  The only place Scheme names appear
in the generated code is in string literals passed to
\texttt{SchemeSymbol.Symbol()} during registration, where they are
harmless.

\subsection{Self-Tail-Call and Let Shadowing}

The serial-numbered scheme interacts cleanly with the self-tail-call
optimization.  In the generated \texttt{LOOP}, the parameter variables
are always the original \texttt{a\_0}, \texttt{a\_1}, etc.  When a
let-bound variable shadows a parameter, it gets a distinct name from
the \texttt{v\_} series, so the tail-call reassignment
\texttt{a\_0~:=~\ldots;\ a\_1~:=~\ldots} always refers to the
correct LOOP variables, even from within a nested let scope that has
shadowed one of the parameters.

\subsection{Define Deduplication}
\label{sec:impl-dedup}

When compiling a large Scheme codebase, it is common for the same
file to be loaded more than once (e.g., a utility library loaded
by multiple modules that are later concatenated).  This produces
duplicate \texttt{define} forms that would cause ``symbol
already defined'' errors in the generated Modula-3.

The compiler applies a deduplication pass before compilation:
\texttt{deduplicate-defines} walks the list of extracted defines
in reverse order, keeping only the \emph{first} occurrence of each
function name (which corresponds to the \emph{last} definition in
the source, preserving Scheme's redefinition semantics).  Duplicate
definitions are silently dropped with a diagnostic message.

When the CSP compiler was first compiled as a single concatenated
file (before per-module compilation was introduced), deduplication
removed 4 duplicate definitions from 867 extracted defines,
reducing the unique count to 863.

%----------------------------------------------------------------------
\section{Performance Results}
\label{sec:impl-perf}
%----------------------------------------------------------------------

The test program includes a benchmark section that compares compiled
and interpreted execution of five workloads.  The benchmarks were
run on ARM64\_DARWIN (Apple M4, macOS) with \texttt{m3\_optimize("T")}
enabled.  The current compiler includes intra-module direct calls,
primitive inlining, inline boolean tests, and numeric unboxing
(all described in Chapter~\ref{ch:optimizations}).

\begin{center}
\begin{tabular}{lccc}
\toprule
\textbf{Benchmark} & \textbf{Compiled} & \textbf{Interpreted} & \textbf{Speedup} \\
\midrule
\texttt{(fibonacci 25)}         &  38 ms    &   629 ms  & 16.5$\times$ \\
\texttt{(fibonacci 30)}         & 421 ms    & 6,978 ms  & 16.6$\times$ \\
\texttt{(loop-sum 10000 0)}     & 0.08 ms   &  23.6 ms  & 314$\times$ \\
\texttt{(ackermann 3 4)}        & 0.97 ms   &  36.7 ms  & 37.8$\times$ \\
\texttt{(lambda-map-sum 10000)} & 4.5 ms    &   120 ms  & 26.6$\times$ \\
\bottomrule
\end{tabular}
\end{center}

\noindent
These numbers reflect the cumulative effect of all implemented
optimizations.  The mutual recursion benchmark provides a
three-way comparison---compiled with direct calls, compiled with
general dispatch (direct-call pointer check defeated by forwarding
lambdas), and fully interpreted:

\begin{center}
\begin{tabular}{lccccc}
\toprule
\textbf{Benchmark} & \textbf{Direct} & \textbf{General} & \textbf{Interp} & \textbf{D/G} & \textbf{D/I} \\
\midrule
\texttt{(mut-f 25)} &  41 ms & 325 ms &   636 ms & 8.0$\times$ & 15.6$\times$ \\
\texttt{(mut-f 28)} & 166 ms & 1,378 ms & 2,713 ms & 8.3$\times$ & 16.3$\times$ \\
\bottomrule
\end{tabular}
\end{center}

\noindent
The D/G column shows the cost of virtual dispatch: the 8$\times$
slowdown from ``Direct'' to ``General'' represents the overhead of
\texttt{NARROW} plus virtual method dispatch at every cross-function
call, which the intra-module direct-call optimization eliminates.

\subsection{Analysis}

The compiled code delivers a \textbf{16--314$\times$ speedup}
depending on the workload.  The variation reflects how much of each
benchmark's work is eliminated by the different optimization layers.

\paragraph{Fibonacci (16.5$\times$).}
The \texttt{fibonacci} benchmark is the purest test of call
overhead.  Each call performs a comparison, two subtractions, two
recursive calls, and one addition---very little per-call work.
The intra-module direct-call optimization eliminates
\texttt{NARROW} and virtual dispatch on the fast path.  Primitive
inlining eliminates the binding lookup for \texttt{+}, \texttt{-},
and \texttt{=}.  The 16.5$\times$ speedup shows that the compiler
eliminates nearly all of the interpreter's per-call overhead.

\paragraph{Loop-sum (314$\times$).}
This benchmark demonstrates the combined effect of self-tail-call
optimization and numeric unboxing.  The function
\texttt{(loop-sum n acc)} iterates $n$ times, adding \texttt{n}
to \texttt{acc} each iteration.  The generated code keeps both
parameters as raw \texttt{LONGREAL} values:

\begin{lstlisting}[language=Modula3]
VAR
  nr_0, nr_1 : LONGREAL;
  v_0, v_1 : LONGREAL;
BEGIN
  nr_0 := NARROW(a_0, SchemeLongReal.T)^;
  nr_1 := NARROW(a_1, SchemeLongReal.T)^;
  LOOP
    IF nr_0 = 0.0d0 THEN
      RETURN SchemeLongReal.FromLR(nr_1);
    ELSE
      v_0 := (nr_0 - 1.0d0);
      v_1 := (nr_1 + nr_0);
      nr_0 := v_0;
      nr_1 := v_1;
    END;
  END
END;
\end{lstlisting}

\noindent
The loop body contains \textbf{zero heap allocations}---no
\texttt{SchemeLongReal.FromLR()} calls, no \texttt{NARROW} calls.
Arithmetic is performed directly on unboxed \texttt{LONGREAL} values.
Only at the return point is a single \texttt{FromLR} allocation
needed to re-box the result.  On ARM64, the loop body compiles to
approximately 5 machine instructions (compare, subtract, add,
two moves, branch).  The 314$\times$ speedup over the interpreter
reflects the elimination of \emph{all} overhead: no eval loop, no
type dispatch, no call overhead, no boxing.

\paragraph{Ackermann (37.8$\times$).}
The Ackermann function involves
deeply nested non-tail recursion with very small bodies.
Direct calls and primitive inlining eliminate nearly all per-call
overhead, yielding a 37.8$\times$ speedup.

\paragraph{Lambda-map-sum (26.6$\times$).}
This benchmark exercises lambda closures, higher-order calls, and
list traversal.  The 26.6$\times$ speedup shows that closure
allocation, free-variable capture, and the \texttt{apply} dispatch
are now dominated by the actual list operations rather than
call overhead.

\subsection{Remaining Per-Call Overhead}
\label{sec:impl-bottleneck}

For calls to functions \emph{outside} the compilation unit (e.g.,
functions from other packages, or interpreted functions), the
per-call overhead remains:

\begin{enumerate}
\item \texttt{self.b\_\emph{foo}.get()} --- one binding lookup
  (pointer dereference, array index, load) to fetch the callee.
\item \texttt{NARROW(\emph{proc}, SchemeProcedure.T)} --- a runtime
  type check.
\item \texttt{.apply1(interp, arg)} --- a virtual method dispatch
  (indirect call through the method table).
\end{enumerate}

\noindent
For intra-module calls, the direct-call optimization
(\S\ref{sec:opt-direct}) eliminates steps~2 and~3 on the fast
path, reducing the overhead to a single binding lookup plus a
well-predicted branch.  The self-tail-call optimization eliminates
all three steps entirely for tail-recursive functions.

%----------------------------------------------------------------------
\section{SchemeCompiledRegistry}
\label{sec:impl-registry}
%----------------------------------------------------------------------

The \texttt{SchemeCompiledRegistry} module provides transparent
loading of compiled Scheme modules.  When an MScheme application
calls \texttt{(load "foo.scm")}, the runtime checks the registry
before falling back to interpreted loading.  If a compiled module
named \texttt{foo} has been registered (via its \texttt{Install}
procedure at startup), the compiled definitions are used instead
of parsing and interpreting the source file.

This means that existing Scheme code that uses \texttt{(load ...)}
or \texttt{(require-modules ...)} transparently picks up compiled
versions without any source changes.  The CSP compiler, for example,
loads $\sim$50 Scheme modules via \texttt{(load "setup.scm")} chains;
with \texttt{SchemeCompiledRegistry}, 49 of these are served from
compiled code.

%----------------------------------------------------------------------
\section{Large-Scale Validation: The CSP Compiler}
\label{sec:impl-cspc}
%----------------------------------------------------------------------

The unit test suite (\S\ref{sec:impl-tests}) validates correctness on
isolated functions.  To validate the compiler on a real, large-scale
Scheme codebase, we compiled the entire CSP (Communicating Sequential
Processes) compiler from the Intel Labs async-toolkit project.  The
CSP compiler (\texttt{cspc}) is a production tool for asynchronous
circuit design, written in MScheme, totaling approximately 17,000
lines of Scheme across $\sim$50 source files.

\subsection{Compilation Results}

The $\sim$50 source files are compiled individually as separate
modules during the \texttt{cm3} build.  Across all modules, the
compiler processes approximately 1,236 top-level \texttt{define}
forms.  Of these:

\begin{center}
\begin{tabular}{lr}
\toprule
\textbf{Category} & \textbf{Count} \\
\midrule
Successfully compiled    & 724 (59\%) \\
Skipped (passthrough to interpreter) & 512 (41\%) \\
\midrule
Total defines     & $\sim$1,236 \\
\bottomrule
\end{tabular}
\end{center}

The remaining skipped definitions are blocked primarily by
\texttt{quasiquote} (used in template-based code generation)
and by patterns not yet handled by the compilability checker
(e.g., complex \texttt{apply} patterns, higher-order functions
that take variable numbers of procedure arguments).
Key modules achieve high compilation ratios:
\texttt{simplify} (7/7), \texttt{expr} (15/15),
\texttt{symbolic} (26/27), \texttt{product} (12/13),
\texttt{visit} (12/13), \texttt{higher\_order} (23/23),
\texttt{saturation} (9/10).

\subsection{Correctness Verification}

To verify that the compiled definitions produce identical results
to the interpreted originals, the compiled CSP compiler was run on
all seven MiniMIPS processor designs (MemStage, DataMem, Decode,
Fetch, Execute, RegFile, Writeback).  These are real circuit
specifications that exercise the full depth of the CSP compiler's
analysis, transformation, and code generation passes.

The output of the compiled configuration was compared against the
output of the fully interpreted configuration (all definitions
interpreted via \texttt{MSCHEME\_INTERPRETED=1}).
The results were \textbf{byte-for-byte identical} on all seven
processes, verified by SHA-1 checksum of all generated Modula-3
source files.

This is strong evidence that the compiler preserves semantics
faithfully: the compiled definitions interoperate correctly
with each other, with the interpreted (passthrough) definitions, and with
the MScheme runtime primitives.

\subsection{End-to-End Performance}
\label{sec:impl-cspc-perf}

To measure the performance impact of compilation on a real workload,
the full MiniMIPS processor design (seven processes: MemStage,
DataMem, Decode, Fetch, Execute, RegFile, Writeback) was compiled
by \texttt{cspc} in two configurations: compiled mode (using
\texttt{SchemeCompiledRegistry} for all available compiled
definitions) and interpreted mode (forcing all definitions through
the interpreter via \texttt{MSCHEME\_INTERPRETED=1}).
The output of both runs was verified to be byte-for-byte identical.

\begin{center}
\begin{tabular}{lcc}
\toprule
\textbf{Benchmark} & \textbf{Compiled} & \textbf{Configuration} \\
\midrule
MiniMIPS (7 processes) & 42.6\,s & All optimizations \\
HelloWorld (1 process) & 0.62\,s & All optimizations \\
\bottomrule
\end{tabular}
\end{center}

\noindent
These timings include all implemented optimizations: intra-module
direct calls, primitive inlining, inline boolean tests, and
numeric unboxing.  The MiniMIPS compilation is dominated by the
Execute process (the ALU), which involves abstract interpretation
over large ASTs with thousands of nodes.

The compiled CSP compiler uses 49 of its 53 source modules
in compiled form (the remaining 4---\texttt{cspbuild}, \texttt{sim},
\texttt{test}, and \texttt{vary}---are build scripts or utilities
not used in the compilation path).  Of the $\sim$1,236 top-level
definitions across all modules, approximately 724 (59\%) are
compiled natively; the remainder fall through to the interpreter.

The overall speedup compared to fully interpreted mode (earlier
measured at 86.5\,s for the same workload) is approximately
2$\times$.  This is lower than the 16--314$\times$ seen in
micro-benchmarks because:

\begin{itemize}
\item The workload is dominated by string operations, hash table
  lookups, and list manipulation---built-in operations that are
  already native code in the interpreter.  The compiler's
  optimizations primarily eliminate \emph{dispatch} overhead
  (type checks, virtual calls), not the cost of the operations
  themselves.

\item Not all functions are compiled.  The 41\% that remain
  interpreted include \texttt{quasiquote}-using code generation
  templates and complex higher-order patterns.

\item Cross-module calls between compiled modules still incur
  binding-lookup overhead (the direct-call optimization is
  intra-module only).
\end{itemize}

\subsection{Bugs Discovered}

The CSP compiler stress test uncovered five additional bugs:
the three original bugs
(\S\ref{sec:bug-bodyless}, \S\ref{sec:bug-short-if},
\S\ref{sec:bug-constants}) from conventional code paths, plus
two bugs exposed during the optimization work:

\begin{itemize}
\item \textbf{The \texttt{gen-self-tail-call} arity mismatch.}
  The numeric unboxing code used a 3-list \texttt{map} over
  \texttt{(params args all-temps)}, which crashed when \texttt{args}
  was shorter than \texttt{params} (e.g., for rest-parameter
  functions).  Fixed by replacing \texttt{map} with explicit loops
  that stop at the shortest list.

\item \textbf{The \texttt{set!}+lambda capture bug}
  (\S\ref{sec:opt-letcell}).  A \texttt{let}-bound variable that
  was both \texttt{set!}-mutated and captured by a lambda was
  captured by value instead of by reference, causing the lambda
  to see the init value permanently.  This was a pre-existing
  limitation exposed when the bootstrap-generated
  \texttt{set\_lib.m3} was regenerated.
\end{itemize}

\noindent
This validates the importance of testing on real codebases in
addition to synthetic test suites.

\subsection{Current Restrictions}

The Level~1 compiler supports a broad subset of Scheme in
compiled definitions:

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Supported} & \textbf{Not yet supported} \\
\midrule
\texttt{if}, \texttt{cond} (with \texttt{else} and \texttt{=>})
  & \texttt{quasiquote} \\
\texttt{let}, \texttt{let*}, named \texttt{let}
  & \texttt{call/cc} \\
\texttt{letrec}, internal \texttt{define}
  & General tail-call optimization \\
\texttt{lambda} (closures with captures)
  & \\
\texttt{case}, \texttt{do}
  & \\
\texttt{begin}, \texttt{and}, \texttt{or}, \texttt{not}
  & \\
\texttt{set!} (local and free variables)
  & \\
\texttt{unwind-protect} (via \texttt{TRY}/\texttt{EXCEPT})
  & \\
\texttt{quote} (all types)
  & \\
Self-tail-call optimization
  & \\
Fixed and rest parameters
  & \\
Free-variable capture (by value and by binding)
  & \\
Constants (numbers, strings, symbols,
  & \\
\quad booleans, chars, null, quoted lists)
  & \\
\bottomrule
\end{tabular}
\end{center}

Definitions using any unsupported construct are silently skipped
by the compilability checker and fall back to the interpreter with
no loss of semantics.  The compiled and interpreted definitions
interoperate transparently through the
\texttt{SchemeProcedure.T} interface (\S\ref{sec:proctype}).
The only construct of practical significance that remains
unsupported is \texttt{quasiquote}, which blocks one definition
in the compiler itself (\texttt{form->string}) and a handful of
definitions in the CSP compiler that use template-based code
generation.

%======================================================================
\chapter{Implemented Language Extensions}
\label{ch:implemented}
%======================================================================

The following language constructs were added to the Level~1 compiler
after the initial release.  Each was a gap identified in the original
design; this chapter documents the implementation approach taken
for each.

%----------------------------------------------------------------------
\section{Closures (\texttt{lambda})}
\label{sec:gap-lambda}
%----------------------------------------------------------------------

Lambda was the single largest gap in the original compiler.
The implementation follows the flat-closure strategy described
in Chapter~\ref{ch:analysis}: each \texttt{lambda} expression
generates a \texttt{SchemeProcedure.T} subclass (named
\texttt{Compiled\_lam\emph{N}}) with typed fields for captured
variables.

Captures are classified into two categories:

\begin{description}
\item[Binding captures (\texttt{b\_\emph{N}} fields)]
  Variables that are free in the enclosing function (accessed via
  \texttt{SchemeEnvironmentBinding.T} cells).  The lambda shares
  the same binding cell reference---mutations are visible in both
  directions.

\item[Value captures (\texttt{cap\_\emph{N}} fields)]
  Variables that are parameters or \texttt{let}-bound locals of
  the enclosing function.  These are captured by value
  (\texttt{SchemeObject.T} snapshot at closure-creation time).
\end{description}

The closure type declaration and \texttt{apply} methods are
accumulated during code generation and emitted after the
enclosing function's procedures, exactly as for top-level
\texttt{define} forms.  Lambdas may nest to arbitrary depth;
each level captures from its immediately enclosing scope.

Adding lambda support raised the compiler's self-compilation
ratio from 38/59 to 93/100 (in combination with the other
constructs described below) and the CSP compiler ratio
from 68\%\ to approximately 59\% of a larger total
(many previously-skipped functions now compile, but the
per-module compilation also discovered additional definitions).

%----------------------------------------------------------------------
\section{Named \texttt{let}}
\label{sec:gap-named-let}
%----------------------------------------------------------------------

Named \texttt{let} is compiled to a Modula-3 \texttt{LOOP} with
parameter reassignment, reusing the self-tail-call machinery.
Tail calls to the loop name become variable reassignment followed
by \texttt{LOOP} continuation.

For named \texttt{let} in non-tail position (where \texttt{RETURN}
cannot be used to exit the loop), the compiler emits a
\texttt{WHILE} loop controlled by a boolean flag variable.  The
result is assigned to a target temporary and the flag is set to
\texttt{FALSE} to exit.

%----------------------------------------------------------------------
\section{\texttt{letrec} and Internal \texttt{define}}
\label{sec:gap-letrec}
%----------------------------------------------------------------------

\texttt{letrec} binds variables to undefined values, then evaluates
the initializers in the scope of all the bindings (allowing mutual
references).  Internal \texttt{define} forms are desugared to
\texttt{letrec} (or \texttt{let} when non-recursive) by hoisting
all defines from the body.

The key implementation challenge is that lambda initializers in a
\texttt{letrec} may capture sibling bindings that have not yet been
assigned.  If the bindings were plain Modula-3 local variables,
capture-by-value would snapshot \texttt{NIL}.  The solution uses
\texttt{LetrecCell}---a lightweight \texttt{SchemeEnvironmentBinding.T}
subtype generated inline in each compiled module:

\begin{lstlisting}[language=Modula3]
TYPE LetrecCell = SchemeEnvironmentBinding.T OBJECT
    val : SchemeObject.T;
  OVERRIDES
    get := LCGet; setB := LCSet; ...
  END;
\end{lstlisting}

Each \texttt{letrec}-bound variable is allocated as a
\texttt{LetrecCell} initialized to \texttt{NIL}.  The cells are
added to the compiler's free-variable map, so lambdas in the
init expressions capture them by \emph{reference} (as
\texttt{b\_\emph{N}} binding fields) rather than by value.
After all initializers are evaluated, each cell is set via
\texttt{.setB()}.  When the lambdas are later called, they see the
correct values through the binding cell indirection.

\texttt{LetrecCell.get()} and \texttt{.setB()} are direct field
accesses---no environment hash-table lookup is involved---so the
overhead relative to plain local variables is minimal (one extra
pointer dereference per access).

%----------------------------------------------------------------------
\section{Other Constructs}
\label{sec:gap-other}
%----------------------------------------------------------------------

\paragraph{\texttt{cond} with \texttt{=>}.}
The clause \texttt{(test => proc)} evaluates \texttt{test}; if
truthy, it calls \texttt{proc} with the test value as argument.
The implementation binds the test result to a temporary and
dispatches through \texttt{apply1}.

\paragraph{\texttt{case}.}
Compiled as a chain of \texttt{IF}/\texttt{ELSIF} comparisons
using \texttt{SchemeUtils.Eqv}.

\paragraph{\texttt{do}.}
Compiled to a Modula-3 \texttt{LOOP} with step expressions
evaluated into temporaries and assigned at the end of each
iteration, following the self-tail-call pattern.

\paragraph{Rest parameters.}
Functions with dotted parameter lists
\texttt{(define (f a b .\ rest) ...)} are compiled with the
\texttt{Apply} (list-argument) method extracting fixed parameters
via \texttt{SchemeUtils.First}/\texttt{Rest} and binding the
remainder to the rest parameter.  The \texttt{apply1} and
\texttt{apply2} specializations construct a list from excess
arguments.

\paragraph{\texttt{and}/\texttt{or} in tail position.}
The last argument of \texttt{and} and \texttt{or} is now
recognized as a tail position, allowing named-\texttt{let}
loop calls there.  This unblocked several recursive
predicates in the compiler itself.

%======================================================================
\chapter{Implemented Optimizations}
\label{ch:optimizations}
%======================================================================

This chapter describes the optimizations implemented after the
initial Level~1 compiler.  Together they deliver 16--314$\times$
speedups over the interpreter on micro-benchmarks and a 2$\times$
speedup on the CSP compiler's MiniMIPS workload.

%----------------------------------------------------------------------
\section{Intra-Module Direct Calls}
\label{sec:opt-direct}
%----------------------------------------------------------------------

For calls between functions defined in the same compilation unit,
the compiler emits a pointer comparison that guards a direct
procedure call, bypassing both \texttt{NARROW} and virtual dispatch.

For each function \texttt{foo} in the compilation unit, the
\texttt{Install} procedure stores a typed reference in a
module-level \texttt{VAR}:

\begin{lstlisting}[language=Modula3]
VAR direct_p3 : Compiled_p3;  (* typed ref to foo *)
\end{lstlisting}

\noindent
At each call site, the generated code compares the live binding
against the direct reference:

\begin{lstlisting}[language=Modula3]
VAR fn_temp := self.b_0.get();
BEGIN
  IF fn_temp = direct_p3 THEN
    (* fast path: direct call, no NARROW, no vtable *)
    t_0 := Apply1_p3(direct_p3, interp, a_0);
  ELSE
    (* slow path: general dispatch *)
    t_0 := NARROW(fn_temp,
        SchemeProcedure.T).apply1(interp, a_0);
  END;
END;
\end{lstlisting}

\noindent
\textbf{Redefinition semantics are preserved:} the binding is
always checked first.  If \texttt{foo} has been redefined (the
pointer comparison fails), the slow path handles it correctly.
Modern branch predictors achieve near-100\% accuracy on the fast
path because redefinition is rare.

The mutual recursion benchmark (\S\ref{sec:impl-perf}) shows the
impact: \texttt{(mut-f 25)} runs 8$\times$ faster with direct
calls than with general dispatch, demonstrating that the
\texttt{NARROW} + virtual dispatch overhead dominates for
call-intensive workloads.

%----------------------------------------------------------------------
\section{Primitive Inlining}
\label{sec:opt-inline}
%----------------------------------------------------------------------

Calls to known built-in primitives are inlined directly as
Modula-3 expressions, eliminating the binding lookup, NARROW, and
virtual dispatch entirely.  The compiler maintains a table of
inlineable primitives and their M3 equivalents:

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Scheme} & \textbf{Inlined Modula-3} \\
\midrule
\texttt{(+ a b)}     & \texttt{SchemeNumber.Add(a, b)} \\
\texttt{(- a b)}     & \texttt{SchemeNumber.Sub(a, b)} \\
\texttt{(* a b)}     & \texttt{SchemeNumber.Mul(a, b)} \\
\texttt{(/ a b)}     & \texttt{SchemeNumber.Div(a, b)} \\
\texttt{(car x)}     & \texttt{SchemeUtils.First(x)} \\
\texttt{(null? x)}   & \texttt{SchemeBoolean.FromBool(x = NIL)} \\
\texttt{(pair? x)}   & \texttt{SchemeBoolean.FromBool(ISTYPE(x, SchemePair.T))} \\
\texttt{(number? x)} & \texttt{SchemeBoolean.FromBool(SchemeNumber.Is(x))} \\
\texttt{(cons a b)}  & \texttt{SchemePair.Cons(a, b)} \\
\texttt{(eq? a b)}   & \texttt{SchemeBoolean.FromBool(a = b)} \\
\texttt{(not x)}     & \texttt{SchemeBoolean.FromBool(NOT SchemeBoolean.Truth(x))} \\
\bottomrule
\end{tabular}
\end{center}

\noindent
Arithmetic operations (\texttt{+}, \texttt{-}, \texttt{*},
\texttt{/}) delegate directly to
\texttt{SchemeNumber.Add/Sub/Mul/Div}.  This handles the full numeric
tower (exact integers, rationals, inexact reals, MPFR, complex, duals)
with correct contagion rules.  The \texttt{number?} predicate
compiles to \texttt{SchemeNumber.Is(x)}, recognizing all numeric types.
Rational, complex, and dual constants are embedded via
\texttt{SchemeRational.New}, \texttt{SchemeComplex.T},
and \texttt{SchemeDual.New} respectively.  The generated code imports
\texttt{SchemeNumber}, \texttt{SchemeInt}, \texttt{SchemeRational},
\texttt{SchemeComplex}, and \texttt{SchemeDual} as needed.

Primitives are inlined unconditionally---no guard check is emitted.
Since R4RS reserves the right to make arithmetic operators
syntactic keywords, and primitive redefinition is vanishingly
rare in practice, this is a reasonable trade-off.  If a user
redefines \texttt{+}, the compiled code will still use the
original arithmetic.

%----------------------------------------------------------------------
\section{Inline Boolean Tests}
\label{sec:opt-booltest}
%----------------------------------------------------------------------

When a numeric comparison appears as the test of an \texttt{if}
expression, the compiler emits the comparison directly as a
Modula-3 boolean expression, skipping the \texttt{SchemeBoolean.TruthO()}
dispatch.  For example:

\begin{lstlisting}[language=Scheme]
(if (= n 0) acc (loop (- n 1) (+ acc n)))
\end{lstlisting}

\noindent
Previously compiled as:

\begin{lstlisting}[language=Modula3]
IF SchemeBoolean.TruthO(
     SchemeNumber.NumEq(a_0, self.k_0)) THEN
\end{lstlisting}

\noindent
Now compiled as:

\begin{lstlisting}[language=Modula3]
IF NARROW(a_0, SchemeLongReal.T)^ = 0.0d0 THEN
\end{lstlisting}

\noindent
This eliminates the intermediate \texttt{SchemeBoolean.T} allocation
and the \texttt{TruthO} type-check dispatch.
Note: the \texttt{NARROW}-based inline form currently assumes
\texttt{LONGREAL} operands; non-\texttt{LONGREAL} numeric types
fall back to the \texttt{SchemeNumber} dispatch path.  The optimization
applies to \texttt{=}, \texttt{<}, \texttt{>}, \texttt{<=}, and
\texttt{>=} in test position.  Compound boolean expressions using
\texttt{and}/\texttt{or} are also inlined when their leaves are
inlineable comparisons.

The \texttt{(not expr)} form in test position flips consequent and
alternative rather than emitting a negation.

%----------------------------------------------------------------------
\section{Numeric Unboxing}
\label{sec:opt-unboxing}
%----------------------------------------------------------------------

In self-tail-call loops and named-\texttt{let} loops, parameters
used exclusively in numeric contexts are ``unboxed''---kept as raw
\texttt{LONGREAL} values throughout the loop body, eliminating
all heap allocation per iteration.

\subsection{Analysis}

The function \texttt{analyze-numeric-params} examines the procedure
body and determines which parameters are used exclusively in numeric
contexts.  A parameter is numeric-only if:

\begin{enumerate}
\item Every \textbf{use} of the parameter is in a numeric context:
  argument to an inlineable arithmetic op (\texttt{+}, \texttt{-},
  \texttt{*}, \texttt{quotient}, \texttt{remainder}) or numeric
  comparison (\texttt{=}, \texttt{<}, \texttt{>}, \texttt{<=},
  \texttt{>=}), or passed to a self-tail-call in the same position.

\item Every \textbf{value} assigned to it in self-tail-calls is
  either a numeric constant, an arithmetic expression, or another
  numeric-only parameter.
\end{enumerate}

\noindent
The analysis requires a ``forcing numeric use''---at least one
direct arithmetic operation or numeric comparison on the
parameter---to ensure the parameter is genuinely numeric, not
merely passed through to tail calls.

\subsection{Code Generation}

Unboxed parameters are declared as \texttt{LONGREAL} and
initialized by narrowing the \texttt{SchemeObject.T} arguments
once at procedure entry:

\begin{lstlisting}[language=Modula3]
VAR
  nr_0, nr_1 : LONGREAL;
  v_0, v_1 : LONGREAL;
BEGIN
  nr_0 := NARROW(a_0, SchemeLongReal.T)^;
  nr_1 := NARROW(a_1, SchemeLongReal.T)^;
  LOOP
    IF nr_0 = 0.0d0 THEN
      RETURN SchemeLongReal.FromLR(nr_1);
    ELSE
      v_0 := (nr_0 - 1.0d0);
      v_1 := (nr_1 + nr_0);
      nr_0 := v_0;
      nr_1 := v_1;
    END;
  END
END;
\end{lstlisting}

\noindent
Inside the loop: zero \texttt{FromLR} allocations, zero
\texttt{NARROW} calls.  Arithmetic is performed directly on
\texttt{LONGREAL} values.  At return points, a single
\texttt{SchemeLongReal.FromLR()} re-boxes the result.

\subsection{Safety}

If an unboxed variable is used in a non-numeric context (e.g.,
passed to \texttt{cons} or \texttt{display}), the \texttt{var-ref}
function automatically wraps it with
\texttt{SchemeLongReal.FromLR(nr\_N)}, ensuring correctness at the
cost of one allocation at that use point.  The analysis is
conservative: any non-numeric use prevents unboxing.

\subsection{Mixed Parameters}

Parameters are analyzed independently.  In
\texttt{(define (f n lst) ...)}, where \texttt{n} is used only in
arithmetic and \texttt{lst} is used with \texttt{car}/\texttt{cdr},
only \texttt{n} is unboxed; \texttt{lst} remains a boxed
\texttt{SchemeObject.T}.

\subsection{Current Status}

Numeric unboxing is currently enabled only for \texttt{LONGREAL}
parameters.  With the transition to \texttt{SchemeNumber}-based
arithmetic (supporting exact integers, rationals, complex, and
duals), unboxing for non-\texttt{LONGREAL} types is disabled
pending a type-inference pass that can determine the concrete
numeric type at compile time.

%----------------------------------------------------------------------
\section{The \texttt{set!}+Lambda Capture Fix}
\label{sec:opt-letcell}
%----------------------------------------------------------------------

A correctness fix (not a performance optimization) was required for
\texttt{let}-bound variables that are both mutated via \texttt{set!}
and captured by a \texttt{lambda}.  The pattern:

\begin{lstlisting}[language=Scheme]
(let ((res '()))
  (set! res (lambda (x) ... res ...))
  res)
\end{lstlisting}

\noindent
Previously, \texttt{res} was captured by value (as a
\texttt{cap\_N} field snapshot at closure-creation time).  Since
the \texttt{set!} occurs after the lambda is created, the lambda's
copy was permanently the init value (\texttt{NIL}).

The fix detects \texttt{let}-bound variables that are both
\texttt{set!} targets and lambda-captured, and promotes them to
\texttt{SchemeEnvironmentBinding.T} cells (like \texttt{letrec}).
Three helper functions implement the detection:
\texttt{set!-targets} finds all \texttt{set!} target symbols,
\texttt{lambda-captured-vars} finds variables appearing free inside
lambdas, and \texttt{needs-binding-cell?} returns true when both
conditions hold.  The promoted variables are accessed via
\texttt{.get()} and \texttt{.setB()} like \texttt{letrec} bindings.

This bug was discovered when the CSP compiler's \texttt{set.scm}
module (which uses this pattern in \texttt{make-set}) was
regenerated by the current compiler instead of using the
bootstrap-generated version.

%======================================================================
\chapter{Remaining Compiler Gaps}
\label{ch:gaps}
%======================================================================

With the language extensions described in
Chapter~\ref{ch:implemented} and the optimizations in
Chapter~\ref{ch:optimizations}, the compiler now handles all
commonly-used Scheme constructs and delivers substantial speedups.
The compiler self-compiles with a 3-pass bootstrap reaching
fixed-point (identical MD5 across passes).
Definitions that use unsupported
constructs are silently skipped by the compilability checker
(\S\ref{ch:boundary}), falling back to the interpreter with no
loss of semantics.

This chapter catalogues the remaining gaps.

%----------------------------------------------------------------------
\section{Missing Language Constructs}
\label{sec:gaps-lang}
%----------------------------------------------------------------------

\subsection{\texttt{quasiquote}}
\label{sec:gap-quasiquote}

\paragraph{Impact.}
One definition in the compiler (\texttt{form->string}) uses
\texttt{quasiquote}, and a handful of CSP compiler modules
use it for template-based code generation.

\paragraph{Strategy.}
\texttt{quasiquote} is a reader macro that expands to nested
\texttt{cons}, \texttt{list}, and \texttt{append} calls.
The expansion can be performed as a desugaring pass before
compilation, converting quasiquoted forms to ordinary
\texttt{list}/\texttt{cons}/\texttt{append} expressions
that the compiler already handles.

\paragraph{Difficulty: moderate.}
The expansion rules for nested quasiquote are subtle
(R5RS~\S7.1.4), but the resulting expressions are straightforward
to compile.

\subsection{\texttt{call/cc}}
\label{sec:gap-callcc}

\paragraph{Impact.}
Not used in any current MScheme codebase.

\paragraph{Difficulty: hard.}
Requires continuation capture, which is fundamentally at odds
with the compiled code's use of the native call stack.

%----------------------------------------------------------------------
\section{Call Optimization}
\label{sec:gaps-calls}
%----------------------------------------------------------------------

\subsection{Cross-Module Direct Calls}
\label{sec:gap-cross}

The intra-module direct-call optimization (\S\ref{sec:opt-direct})
requires the caller to hold a typed reference to the callee's
compiled object.  For functions in \emph{different} compilation
units, the compiler does not have such a reference at
code-generation time.

\paragraph{Strategy: registration protocol.}
When module~A calls a function \texttt{bar} defined in module~B,
the fast path requires a typed reference.  This can be provided
through a two-phase initialization: module~B registers its compiled
objects in a well-known table, and module~A looks them up during
its own \texttt{Install}.  The generated code uses the same guard
pattern as intra-module, but with an untyped
\texttt{SchemeProcedure.T} reference (cannot avoid virtual dispatch
across modules, but can skip \texttt{NARROW}).

\paragraph{Difficulty: moderate.}
Requires a cross-module registration protocol and careful ordering
of \texttt{Install} calls.

\subsection{General Tail-Call Optimization}
\label{sec:gap-tco}

The compiler currently optimizes only \emph{self}-tail-calls
(direct recursion).  Mutual tail calls (A tail-calls B, B
tail-calls A) and tail calls to arbitrary functions are not
optimized---each becomes a regular call that consumes stack.

\paragraph{Strategy: trampoline.}
The standard technique for general TCO in a language without
goto is the \emph{trampoline}: instead of calling the tail
callee directly, the function returns a thunk (or a tagged value)
that the trampoline loop invokes.  This eliminates stack growth
at the cost of one allocation per tail call.

\paragraph{Difficulty: hard.}
Requires a new return convention and careful integration with
the interpreter's existing call paths.  Given that self-tail-call
optimization already handles the most common case (accumulator
loops), general TCO is lower priority.

%----------------------------------------------------------------------
\section{Recommended Priorities}
\label{sec:gaps-priorities}
%----------------------------------------------------------------------

The remaining gaps, ordered by expected impact:

\begin{enumerate}
\item \textbf{Cross-module direct calls} --- moderate difficulty,
  would extend the intra-module direct-call benefit across package
  boundaries (\S\ref{sec:gap-cross}).

\item \textbf{\texttt{quasiquote}} --- moderate, would complete
  self-compilation and unblock template-heavy CSP modules
  (\S\ref{sec:gap-quasiquote}).

\item \textbf{General tail-call optimization} --- hard, lowest
  priority since self-tail-calls cover the common case
  (\S\ref{sec:gap-tco}).
\end{enumerate}

%======================================================================
\chapter*{References}
\addcontentsline{toc}{chapter}{References}
%======================================================================

\begin{enumerate}
\item The companion report: \emph{MScheme Internals: A Technical
  Report on the Interpreter Implementation}, March 2026.  Describes
  the binding optimization, evaluator loop, environment system, and
  primitive dispatch in full detail.

\item Peter Norvig, ``JScheme: Scheme in Java,''
  \url{https://norvig.com/jscheme.html}, 1998.
  The origin of the MScheme design.

\item Marc Feeley and Guy Lapalme, ``Using Closures for Code
  Generation,'' \emph{Computer Languages}, 12(1):47--66, 1987.
  Describes the technique of compiling Scheme to closures, which is
  closely related to the approach proposed here.

\item R.\ Kent Dybvig, \emph{Three Implementation Models for Scheme},
  PhD thesis, University of North Carolina at Chapel Hill, 1987.
  Describes heap-based, stack-based, and string-based implementation
  models for Scheme, including closure conversion and environment
  representation strategies relevant to this design.

\item\label{ref:nelson} Greg Nelson (ed.), \emph{Systems Programming with Modula-3},
  Prentice Hall, 1991.

\item Samuel P.\ Harbison, \emph{Modula-3}, Prentice Hall, 1992.

\item Richard Kelsey, William Clinger, and Jonathan Rees (eds.),
  ``Revised$^5$ Report on the Algorithmic Language Scheme,'' 1998.
  \url{https://www.schemers.org/Documents/Standards/R5RS/}.
\end{enumerate}

\end{document}
